In the previous chapters we showed that if the texts about Jesus are mostly accurate, then Jesus must have been a royal figure with extremely deep ties to the Greek and Egyptian traditions of divine rule.
In this scenario, Jesus as the rightful Christos and his companions as highly powerful and educated people.

However, at this point we still have to deeply consider the possibility that Jesus, just like Amon, Ra, Zeus and Dionysus, may have been a mythical figure.
While historically there was a fairly strong consensus Jesus was a real historical figure, more and more scholars notice the arguments for the historicity of Jesus, Paul and the apostles are a lot weaker when investigated carefully.
Lack of contemporary sources, contradictions in the sources, and the extreme similarity of the Jesus story to many other mythical stories of the time, all point to the possibility Jesus was a mythical figure.
The impossibly advanced church structures in the earliest epistles of Paul call his existence into question.

There is a very strong case to be made that the Pauls epistles are all pseudepigraphic, and the gospels could be written as late as 150AD.
And we actually agree, with the current frameworks of biblical scholarship, with christianity starting from zero as a Jewish sect in year 30AD, the most likely dating is indeed mid 2nd century.
If that was the case, then Jesus was most likely a mythical figure, and the gospels were written as a part of the greek imperial propaganda.

However, within the framework of Jesus being a royal figure, coming into the existing greek and egyptian royal cult and ecclesial government structures, the dating of the gospels can be revisited.
Indeed if all evidence is considered more carefully, we actually find the literary evidence was most likely written much earlier than anyone thought possible, not unlikely to the times of Jesus himself.

Importantly, while the gospels are treated as the earliest and most important sources about Jesus, we have to be aware that the gospels certainly had many editions.
There is a pretty strong consensus that what we know as John is the third edition, without ending chapter stating the beloved disciple is getting old.
Luke could have had an edition before the birth narrative which may be what we know as Marcion's gospel.
Mark is known to have an edition without the resurrection.

Moreover with the dating of the gospels, we can also revisit the dating of the large number of early apocryphal texts.
The apocryphal texts were historically dismissed as late developments, but bundling all of them together under one term was obviously a giant mistake of past scholars causing massive bias.
The large majority of the apocryphal texts are very late and clearly written as derivative texts on pre-existing material with no historical value.
However, there are still many texts that demonstrably are very early, and in some cases even earlier than the gospels.
Gospel of Thomas is quite likely predating the four canonical gospels.
The Protoevangelium of James shaped Catholic and Orthodox Marian theology; its narratives of Mary's childhood, perpetual virginity, and the birth of Jesus entered liturgy and iconography even though the text remained non-canonical.
While indeed the church fathers mentioned a list of scripture that constituted the bible today, nobody took much of a note for fifteen centuries and Protovangelion of James was copied just as often as the gospels.
It truly fell out of favor after the Council of Trent in mid 16th century, when the catholic church formally canonized the bible as we know it today.

When we realize there was a very large number of texts written immediately after the events of Jesus, written by different authors, and many of them were likely written by eyewitnesses, with editions reflecting prominent events like the crucifixion of Peter and the destruction of the temple, the chances of this story being a myth fall to near zero.

\section{John Gospel}\label{sec:john-gospel}

\subsection{Dating of the Gospel of John}\label{subsec:dating-of-the-gospel-of-john}

Famously gospel of John is attested the earliest and has the most prolific earliest sources, yet it is dated the latest, purely based on bias coming from invalid assumptions about ``christology''.
This is one of these 19th century assumptions of early biblical scholars that falls apart completely when under any detailed scrutiny, but remains a dogma of the modern biblical scholarship.

By attested, we have a papyrus P52 of John, which is dated to 125AD, plausibly even earlier to 100AD .
The allegedly oldest Mark, Matthew and Luke are attested by Irenaeus in 180AD together with several other gospels that are deemed by Irenaeus as heretical, and the earliest manuscript of Mark (P45) is dated to the early 3rd century.
Although the oldest mention of Marcion is from Tertullian, Tertullian himself gives credible attestation of Marcion in 140AD.
So, by current manuscript and reference chronology, the order of attestation is: John first, followed by Marcion, then Mary Magdalene, Thomas, Judas, and only later Mark, Matthew, and Luke.

This alone can be explained by the accident of history, most early sources were lost, and the oldest was John, just by accident.
However, we need to consider the broader Egyptian manuscript context.
P52 is the oldest manuscript we have of any gospel.
Its exact findspot is unknown, but it is clearly Egyptian; palaeographic dating places it anywhere from the early first century to mid second century.
The Oxyrhynchus papyri---a large Egyptian collection with documented provenance---include gospel fragments alongside dated tax documents, and the John fragments there are among the earliest.

Although the dating can be disputed, on the basis of geography, in Oxyrhynchus, Egypt, we have gospels that were written likely in the same room by the same group of people, so the relative dating of the gospels as they arrived to Oxyrhynchus is rather certain.

Certainly, it could be the case that the synoptic gospels arrived to Oxyrhynchus much later just by accident.
The Nag Hammadi library, dated to the mid second century, preserves no synoptic gospels at all---only the Gospel of Thomas, Gospel of Philip, Gospel of Truth, and other texts that show heavy Johannine influence.
The Apocryphon of John (a gnostic elaboration) appears in multiple copies, indicating that Johannine material circulated widely in Egypt before the synoptics arrived.

The third source fairly securely dated to second century and likely early second century is the Bodmer papyrus, P66 with the Gospel of John.
As in the other cases, the Gospel of John pre-dates the synoptic Gospels in the Bodmer collection by at least 25-50 years.

This is essentially it for Gospel references pre-140AD.
The only possible earlier references to Gospels are Papias, Polycarp and Clement, who plausibly wrote before 140AD.
In case of Papias, he refers to gospels not written in Greek, so they are almost certainly not any of the canonical gospels nor any of the widely known apocryphal gospels.
Clement and Polycarp are quoting Jesus as in the Gospels, but regardless of whether Jesus really did say the words, there is no indication whether they quoted the gospels or the gospels from them or mostl likely they simply quoted the same text or oral tradition.
The next reference is Justin Martyr, and he again is most likely quoting the Gospel of John, notably not any quote of Jesus recorded in the Gospel of John, but a known fragment of John.

Then the later references from Tertullian, Irenaeus, and Origen all contain unambiguous references to all the canonical gospels as already mature and wide-spread texts, but at this point we are already in the year 180AD and later.

To be very clear, this does not indicate John was written in 125AD, nor that Marcion was written in 140AD .
John may have been written in right after the events described, or it may have been written in 100AD .

What that does show is that based on the dates of manuscripts and references, it should be John to have the highest probability of being the earliest gospel, and not Mark, Matthew or Luke.

The author of the gospel of John is also the author of the three epistles of John based on the writing style and vocabulary.
While Eusebius attributes the Book of Revelation to John, the same analysis shows the author of the gospel of John most definitely not the author of the Book of Revelation.
Hence, the John of Patmos who is the only overt author of a new testament book, is not the same person as the author of the gospel of John.
Why this matters is that the author of the John epistles is also heavily emphasizing that he is an eyewitness, who heard, saw and touched Jesus.

One common argument against the early dating of John is the speeches of Jesus are too literary and too long for an oral tradition.
This is most certainly true, but the mainstream conclusion that it means the speeches were all made up and must have been up to 100 years later is simply not logical at all.
A far more sensible conclusion is that the speeches already existed in written form when Jesus gave them.
The apparent references in John of the sayings from the Gospel of Thomas can actually point to not John and Thomas arguing about theology, but rather both using the same source as Jesus.
As pointed out, Jesus must have been extremely well read, and so it is to be expected that he would at least some notes written down by either him or his followers such as Thomas.

There is a number of arguments about dating John past year 60AD based on the last chapter, John 21.
That chapter seems to be written by a different author and is overtly an appendix to the original text that ends at chapter 20 with an obvious ending ``
Now Jesus did many other signs in the presence of the disciples, which are not written in this book;
but these are written so that you may believe that Jesus is the Christ, the Son of God, and that by believing you may have life in his name.
``.
That is making the dating of John's chapters 1-20 quite firmly substantially before 60AD .

Once the arguments from chapter 21 are removed together with high Christology arguments, all other arguments center on Christian - Jewish split.
As is argued in this book, the Christian - Jewish split is a big misconception and never happened.
Hence there are actually no valuable arguments for dating John past 60AD .

\section{The Identity of the Beloved Disciple and the Eyewitness Voice}\label{sec:identity-beloved-disciple}

\subsection{The Gospel of John as Eyewitness Testimony: Mary Magdalene, Not John son of Zebedee}\label{subsec:the-gospel-of-john-is-widely-accepted-to-be-one-gospel-that-indicates-it-was-written-by-an-eyewitness.}

The Gospel of John stands apart from the Synoptics in its persistent claim to be based on the direct testimony of an eyewitness: ``the disciple whom Jesus loved.'' While tradition has long identified this figure as John, son of Zebedee, the internal evidence tells a different story. Unlike Mark, Luke, and Matthew---none of which claim firsthand authorship---John's narrative is saturated with details only a present, emotionally invested observer could recall: private dialogues, household scenes, and the inner thoughts of Jesus' closest companions. Yet, the text never names the Beloved Disciple as John, nor does it place him among the Twelve in a straightforward way. Instead, the pattern of presence at the cross, the tomb, and the resurrection, as well as the unique access to the women around Jesus, all point to Mary Magdalene as the most plausible eyewitness behind the text. Her centrality in the resurrection narratives, her unique commission by Jesus, and her consistent appearance when the male disciples are absent or fleeing, make her the best candidate for the Beloved Disciple. The attribution to ``John'' is best understood as a later convention, perhaps reflecting a redactor or the name of a later scribe, but the substance of the testimony is unmistakably that of Mary Magdalene---whose voice and memory shaped the core of the Fourth Gospel. This explains both the vividness of the narrative and its divergence from the Synoptics, and situates Mary Magdalene as the primary eyewitness behind the text.

The argument operates on two levels.
First, the internal evidence of the Gospel of John points to a female author: the sensory register, the domestic topography of its scenes, the recurring centering of women at narrative hinges, and the emotional vocabulary all belong to a gendered literary pattern well attested in Greek and Egyptian sources.
This claim stands independently of any particular candidate.
Only after establishing that the author was most likely a woman can we turn to the second question: which woman among the known circle of Jesus was capable of writing such a text.
Mary Magdalene is the strongest candidate, but she is not the only one, and the evidence for her specific identity is inferential rather than structural, drawing on early Christian texts outside the later canon rather than internal literary markers.

Archaeology has repeatedly confirmed John's detailed Jerusalem topography, strengthening the case for eyewitness testimony.
The Pool of Bethesda with its five porticoes (John 5:2) was discovered in the nineteenth century exactly where John placed it, with the distinctive five-colonnade structure matching the text.
The Pool of Siloam (John 9:7) was excavated in 2004 and confirmed as a massive Second Temple period ritual pool.
John names the Kidron valley, Gabbatha (the Stone Pavement), Jacob's well, Bethany beyond the Jordan, and Ephraim near the wilderness---all sites that fit pre-70 CE Jerusalem and its environs with a precision no other gospel matches.
John's Galilean details are equally accurate: the relation of Cana to Capernaum, the note about ``going down'' to lakeside towns, and the distances between sites all match actual elevations and geography.
Of all four gospels, John preserves the most precise micro-topography of Jerusalem and Judea, exactly what we would expect from an eyewitness who knew the city intimately before its destruction in 70 CE.

\subsection{ὁ μαθητὴς ὃν ἠγάπα ὁ Ἰησοῦς}\label{subsec:ux1f41-ux3bcux3b1ux3b8ux3b7ux3c4ux1f74ux3c2-ux1f43ux3bd-ux1f20ux3b3ux3acux3c0ux3b1-ux1f41-ux1f30ux3b7ux3c3ux3bfux1fe6ux3c2}

The frequent reference to ``the disciple whom Jesus loved'' may indicate this disciple was not one of the Twelve.
John names Peter, Andrew, Philip, Nathanael, Thomas, both Judases, and in chapter 21 the sons of Zebedee---yet the Beloved Disciple is never identified as any of them.
If the Beloved Disciple were one of the named Twelve, it would have been natural to say so.
The disciple whom Jesus loved seems separate from the Peter's group.
Most notably all gospels seem to strongly indicate all male apostles fled after the arrest of Jesus, and the only ones who stayed were women.
``There were also women looking on from a distance\ldots{} Mary Magdalene, Mary the mother of James\ldots{} and many other women\ldots'' This means we are faced with either a very major contradiction between John and all three Synoptics where there was no reason to introduce one, or --- far more plausibly --- the Beloved Disciple was not one of the Twelve, and was a woman, preserved anonymously but recognized within the early community as an authoritative witness.

\subsection{Γύναι, ἰδοὺ ὁ υἱός σου}\label{subsec:ux3b3ux3cdux3b9-ux1f30ux3b4ux3bfux1f7a-ux1f41-ux3c5ux1f31ux3ccux3c2-ux3c3ux3bfux3c5}

The words of Jesus from the cross---“Woman, behold your son” (John 19:26)---have long been interpreted as directed to the Beloved Disciple, yet the ambiguity of the phrase has fueled debate. While some have proposed Joanna, a prominent woman from Herod’s court and a key supporter of Jesus (Luke 8:3), as a plausible candidate for the eyewitness behind John, the cumulative evidence consistently returns to Mary Magdalene. Joanna’s status and access to the elite make her an attractive intermediary or redactional figure, and her connections (possibly even to Junia, “foremost among the apostles”) cannot be dismissed. However, the narrative focus of John---especially at the cross and the tomb---remains squarely on Mary Magdalene. Her presence at every critical moment, her emotional proximity to Jesus, and her unique role in the resurrection account make her the best explanation for the eyewitness perspective of the Gospel. While Joanna remains plausible as a conduit or later transmitter of the tradition, the internal and external evidence overwhelmingly points to Mary Magdalene as the foundational witness whose testimony undergirds the text.

\subsection{Peter vs. Mary Magdalene: Early Testimony and the Beloved Disciple}\label{subsec:peter-vs-mary-magdalene-early-testimony}

When early Christian texts outside the later canon such as the Gospel of Mary address the question of the Beloved Disciple's identity, the central rivalry is not between John and Mary, but between Peter and Mary Magdalene. These texts consistently portray Mary Magdalene as the recipient of special revelation and as the disciple closest to Jesus, often in direct opposition to Peter's authority. The absence of any tradition in which John son of Zebedee is pitted against Mary for this status is telling; the earliest controversies focus on whether Peter or Mary is the rightful interpreter and leader. This pattern further supports the conclusion that Mary Magdalene, not John, was remembered in the earliest circles as the privileged eyewitness and authoritative voice behind the Gospel.

In the Gospel of Mary (Coptic 17--18), Peter complains: ``Did he really speak with a woman in private, without our knowledge? Are we to turn and all listen to her? Did he prefer her to us?''
Levi (Matthew) rebukes him: ``Peter, you have always been hot-tempered\ldots{} If the Savior made her worthy, who are you to reject her? Surely the Savior knows her very well. That is why he loved her more than us.''
The Gospel of Philip preserves another version: ``The companion of the [Savior is] Mary Magdalene. [He loved] her more than [all] the disciples, and used to kiss her often on [the\ldots].''
In Pistis Sophia, Mary asks thirty-nine of the sixty-four questions; Jesus calls her ``blessed,'' ``more blessed than all women on earth''---and Peter complains that she will not stop talking.

These are three independent texts, from different authors, different languages, and likely different communities, yet all preserve the same formula: Jesus loved Mary more than the others.
This is a stable line across early literature.

The Johannine formula ὁ μαθητὴς ὃν ἠγάπα ὁ Ἰησοῦς (``the disciple whom Jesus loved'') is strange if this is simply John son of Zebedee---he could have been named without difficulty.
But the ``beloved more than the others'' motif is repeatedly attached to Mary Magdalene in non-canonical material.
Canonical John and non-canonical Mary/Philip/Pistis Sophia are therefore two different ways of handling the same underlying memory: that there was a disciple whom Jesus loved ``more than us,'' and that many circles knew this disciple as Mary Magdalene.

As Esther de Boer notes, in Gospel of Mary, Gospel of Thomas, and Pistis Sophia it is always Peter who attacks Mary because she is a woman and because she received special revelation.
No early text preserves John son of Zebedee versus Peter rivalry over authority.
All early rivalry traditions are Peter versus Mary---the only candidate matching the Johannine pattern.

\section{Gendered Structures and Literary Technique in John}\label{sec:gendered-structures-literary}

\subsection{Finally, a woman writer would have been very likely to describe themselves based on the feelings towards others, not many male authors write like this.}\label{subsec:finally-a-woman-writer-would-have-been-very-likely-to-describe-themselves-based-on-the-feelings-towards-others-not-many-male-authors-write-like-this.}

The narrator describes relationships through emotional vocabulary: ``Jesus loved Lazarus,'' ``the disciple whom Jesus loved.''
This relational self-identification---defining oneself through the feelings of others rather than through lineage, office, or deed---matches patterns in women's letters from Roman Egypt, where identity is expressed through bonds rather than status.

\subsection{Gender Pairing in John: Measurable Structural Evidence}\label{subsec:jesus-interactions-towards-nicodemus-are-also-very-feminine-close-and-personal-jesus-is-talking-about-being-born-again.}

The Gospel of John pairs male and female interlocutors in a systematic structural pattern, and in every case the woman outperforms the man in understanding, initiative, and theological clarity.
The clearest example is the back-to-back pairing of Nicodemus in John 3 and the Samaritan woman in John 4.

Nicodemus appears as a male Jewish ruler approaching Jesus at night.
He speaks only three times in the entire exchange: John 3:2, 3:4, and 3:9.
After verse 9 he disappears from the dialogue entirely, while Jesus continues in monologue from 3:10 onward.
The Nicodemus scene contains no successful comprehension, no confession, no act of witness, and no narrative momentum generated by Nicodemus himself.

By contrast, the Samaritan woman in John 4 receives the longest sustained one-on-one dialogue Jesus has with any individual in the Gospels.
The direct conversation extends across approximately twenty consecutive verses (John 4:7--26), with multiple speaking turns on both sides.
She engages Jesus in layered theological discussion, asks probing questions, recognizes him progressively, and finally becomes the Gospel's first public evangelist.
The narrator explicitly states that ``many Samaritans believed because of the woman's testimony'' (John 4:39).

These contrasts are objective and measurable:

\begin{itemize}
\item Nicodemus: 3 speaking turns (3:2, 4, 9); dialogue collapses into misunderstanding and withdrawal.
\item Samaritan woman: $\sim$20 verses of sustained dialogue (4:7--26); successful recognition and public proclamation (4:28--29, 39).
\item Outcome: the male elite figure fails to grasp basic imagery; the socially marginal woman grasps Christological truth and spreads it.
\end{itemize}

This is not interpretation but structural architecture.
John allocates vast narrative space, theological progression, and evangelistic success to a woman, while the corresponding male figure receives minimal dialogue and exits in confusion.
Such inversion of the usual pattern of ancient Mediterranean literature---where women typically occupy supporting or negative roles---is highly unusual.
It fits cleanly with the hypothesis that John was authored by a woman whose lived experience and priorities shaped the narrative.

\subsection{The Samaritan Woman (John 4:1-42)---An Intimate Encounter at the Well}\label{subsec:the-samaritan-woman-john-41-42-an-intimate-encounter-at-the-well}

Jesus meets her alone at a well, a setting often symbolizing romantic or covenant relationships in the Old Testament (e.g., Isaac and Rebekah, Jacob and Rachel).
The conversation is deeply personal because Jesus reveals he knows her past (five husbands) yet does not condemn her.
He gradually leads her to recognize him as the Messiah, and she becomes a key figure in spreading his message.
The emotional shift---from skepticism to joy---makes this one of the most transformative personal encounters in the Gospels.

\subsection{The detailed description of the passion seems to add credibility to the claim that the gospel was written by an eyewitness, or at least someone portraying themselves as an eyewitness.}\label{subsec:the-detailed-description-of-the-passion-seems-to-add-credibility-to-the-claim-that-the-gospel-was-written-by-an-eyewitness-or-at-least-someone-portraying-themselves-as-an-eyewitness.}

The gospels seem to indicate the apostles fled and for a good reason, as they were likely rightfully scared to be accused of supporting Jesus in his alleged crime against the Roman state.
It is far more likely though women would have been allowed to stay and not prosecuted by the Romans, who were from a far more patriarchal society and would be very unlikely to conceive of a woman being a serious threat to the state.

\subsection{Mary Magdalene (John 20:11-18)---A Moment of Pure Devotion}\label{subsec:mary-magdalene-john-2011-18-a-moment-of-pure-devotion}

Mary is weeping alone at the empty tomb, showing deep grief.
Jesus appears, but she doesn't recognize him at first, thinking he's the gardener.
The moment becomes deeply personal when Jesus simply says her name, ``Mary'', and she immediately recognizes him.
She calls him ῥαββωνί (Teacher), showing a deeply personal bond.
He then entrusts her with the first announcement of the Resurrection, making her the first witness of Easter.

The emotional texture of this scene---solitary weeping, the intimate ῥαββωνί, the physical reaching---reflects a pattern where women express attraction to male authority figures through emotional intimacy.
Male disciples in ancient sources express devotion through imitation, public defense, and philosophical succession, not through this register.
The scene reads as written from inside female experience of a male teacher.

\subsection{The Bridegroom Motif}\label{subsec:the-bridegroom-motif}

John the Baptist calls Jesus the bridegroom (John 3:29).
The first miracle is at a wedding (John 2), which is unusual as an opening scene for a Gospel.
The idea of Jesus as a bridegroom has strong symbolic meaning in Jewish and early Christian traditions, sometimes linking to the idea of a divine marriage (God and Israel, Christ and the Church).
Some Gnostic and esoteric traditions later emphasized Mary Magdalene's connection to Jesus, seeing her as part of this bridal imagery.

Across all major literary traditions, men write scenes that include weddings, but they do not build narratives around weddings.
From Homer to Shakespeare to modern film, male authors use weddings as comic finales, plot devices, or social rituals, never as the structural spine of a story.
Narratives saturated with bridal imagery, bridegroom vocabulary, marital symbolism, and relational intimacy overwhelmingly come from female storytellers.
John's Gospel belongs to this pattern: it is a wedding-framed, marriage-coded narrative whose thematic architecture matches a female literary instinct rather than any known male tradition.

\subsection{Olfactory Memory as Authorial Fingerprint: John 12:3 and the Fragrance of the Room}\label{subsec:the-anointing-of-jesus-by-a-woman-with-expensive-perfume-john-123-mark-143-9}

John 12:3 contains a sensory detail that is absent from all three Synoptic parallels.
In Mark 14:3--9, Matthew 26:6--13, and Luke 7:36--50, a woman anoints Jesus with perfume, but none of these accounts describe the smell of the room.
Only John writes:

\begin{quote}
``The house was filled with the fragrance of the perfume.''
\end{quote}

A concordance check confirms that this is the only narrative description in the entire New Testament of what an interior space actually smelled like.
The other Gospels include no room-smell descriptions.
Acts includes no such descriptions.
Paul uses fragrance language only metaphorically (e.g., 2 Cor 2:14--16; Eph 5:2; Phil 4:18), never as concrete sensory narration.
Revelation uses incense imagery symbolically (e.g. Rev 8:3--4), not as atmospheric description of a physical space.

John stands alone in providing a sensory, domestic, room-filling olfactory detail.

This detail is also paired with the other explicit olfactory moment in the Gospels: Martha's warning at Lazarus's tomb, ``Already there is a stench'' (John 11:39).
These are the two longest olfactory references in the canonical Gospels, and both occur in scenes driven by women's voices or actions (Mary of Bethany in John 12; Martha in John 11).

Such detail corresponds with known patterns in documentary papyri: women's letters from Greco--Roman Egypt consistently display preoccupation with domestic space, household management, food, clothing, and the bodies of family members, far more than is typical in male civic or philosophical writing.
The olfactory specificity of John 12 and John 11 fits within that pattern of domestic, embodied attention.

These two sensory moments are therefore more than atmospheric color.
They are the only two extended smell descriptions in the Gospels, both attached to women, both linked to domestic or bodily contexts, and both embedded at critical narrative transitions (burial preparation in John 12; resurrection preparation in John 11).
It is materially verifiable evidence of a narrative voice comfortable with domestic interiority and physical detail---traits that align naturally with a female authorial hand.

\subsection{The Domestic-Interior Voice: Classical Parallels}\label{subsec:domestic-interior-voice}

In Greek and Roman literature written by men, domestic settings are functional, not described.
Household spaces are rarely evoked unless for moralizing or satire.
Sensory description is sparse.

John describes the smell of perfume filling a house (12:3), concern about the smell of decomposition (11:39), detailed household scenes at Bethany and Cana, intimate interior settings like the upper room recline, and relational gestures with physical touch and emotional tears.

This is the genre signature of female authorship in antiquity.
Sappho's fragments emphasize domestic space and sensory experience.
Corinna's poetry centers female-occupied space.
Theocritus' ``Women at the Adonis Festival'' (Idyll 15) depicts domestic interior with smell and texture from a female viewpoint.
The Hellenistic epistolary papyri from Oxyrhynchus show women's letters consistently focus on household, food, clothing, and the bodies of family members.

John's sensory pattern aligns with female-authored genres, not male biography.

\subsection{Gendered Focalization and Recognition Scenes}\label{subsec:gendered-focalization}

In Greek rhetorical tradition, male authors foreground action, public honor, speeches, genealogy, and military or forensic events.
Female authors or female-voiced narratives foreground interior emotional states, private pain, relational dynamics, and personal revelation.

John's Gospel is unique among ancient biographies for emphasizing tears (Mary Magdalene weeping, Jesus weeping), focusing on whispered conversations (Nicodemus at night), prolonged emotional arcs (Mary of Bethany's anointing and Jesus' reaction), recognition through a single personal word (``Mary!''), and intimacy with death and grieving (Lazarus).

The recognition scene in John 20 is particularly striking.
Greek narrative theory (Konstan, Burrus) notes that recognition scenes involving name-calling are disproportionately used in female-authored or female-voiced narratives.
John 20's ``Mary!'' followed by ``Rabbouni!'' is the most intimate anagnorisis scene in the entire New Testament.

John's narrator behaves like someone writing from inside emotional experience, not from outside it.
This is a female-coded narrative signature.

\section{Mary Magdalene and the Johannine Theology of Recognition}\label{sec:johannine-theology-recognition}

\subsection{Jewish Wisdom Tradition and the Prologue}\label{subsec:jewish-wisdom-tradition}

Jewish literature includes female narrators of revelation: Daughter Zion laments in Lamentations, Wisdom/Sophia speaking in first person in Proverbs 8 and Wisdom of Solomon, Deborah in Judges 5, Miriam in Exodus tradition.
Female-voiced revelation and female prophetic authority were not alien to Jewish thought.

John's Prologue fits into this tradition.
The Logos imagery echoes Jewish Wisdom/Sophia, not male Roman biography.
Philo's Logos bridges Jewish Wisdom, Greek philosophy, and Temple Shekinah.
If the author were female and educated in Alexandrian Judaism, she would be perfectly positioned to combine feminine Wisdom traditions with Hellenistic philosophical language.
John's Prologue fits this exact pattern.

\subsection{Women's Memory Books}\label{subsec:womens-memory-books}

Papyrus archives from Roman Egypt show that women kept memorial books, preserved recollections of births, deaths, and illnesses, recorded emotional and spiritual experiences, and preserved their husbands' or sons' sayings.

This is the genre John resembles: a memory-narrative with domestic detail, sensory texture, and theological reflection woven into recollection.
John reads like a woman's memory-book, not a male bios.

\subsection{Feminine Syntax and Emotional Vocabulary}\label{subsec:feminine-syntax}

Female-voiced texts in antiquity, especially the papyrus letters from Roman Egypt, favor perception and emotion verbs: seeing, knowing, believing, fearing, weeping, loving.
Their language is interior, relational, and grounded in lived experience.

John's Gospel shares this profile.
The narrator uses perception verbs more than any New Testament author.
The Gospel employs the full emotional lexicon---Jesus troubled, moved, weeping---and pivots on verbs of love (ἀγαπᾶν, φιλεῖν).

This linguistic pattern resembles the voice of women's household letters far more than the elevated prose of male elite writers.
John's syntax and emotional vocabulary align with female-authored narrative in antiquity.

\subsection{Song of Songs and the Garden Encounter}\label{subsec:song-of-songs}

John 20's garden encounter between Mary and the risen Jesus echoes the Song of Songs.
Mary seeks the beloved, weeping, asking the gardener where he is.
Recognition comes when the beloved calls her by name.
Patristic and medieval commentators regularly interpret the Song of Songs Bride as Mary Magdalene or the Church.
Modern scholars like André Feuillet make the John 20--Song of Songs connection explicit.

In Song of Songs, the beloved is female; the one whose longing, searching, and voice frame the narrative.
John's ``disciple whom Jesus loved'' language sits inside this biblical matrix where God/Christ is bridegroom and the beloved is the feminine partner.
The canonical intertext for ``beloved'' is Song of Songs, and in that template the beloved is always feminine.

\subsection{Lazarus and the Bethany Circle}\label{subsec:lazarus-bethany}

A minority but non-fringe scholarly view holds that the Beloved Disciple is Lazarus.
John 11:3 says ``Lord, he whom you love is ill.''
John 11:5 says ``Jesus loved Martha and her sister and Lazarus.''
John 11:36 says ``See how he loved him.''
The same verb appears in John 13, 19, 20, 21 for the disciple whom Jesus loved.

If Lazarus is the beloved disciple, whose narrative vantage point explains the Bethany scenes, the domestic detail, and the private knowledge of Lazarus' illness, death, and raising?
The obvious candidate is Mary of Bethany, the woman closest to Lazarus.
Even if reviewers balk at Mary as the Beloved Disciple, serious scholars already detach the beloved disciple from John son of Zebedee and root the ``beloved'' motif in the Bethany circle, where Mary is central.

\subsection{The Cross and the Editorial Formula}\label{subsec:cross-editorial-formula}

John 19:25 lists four women at the cross, including Mary Magdalene.
In verses 26--27, Jesus speaks to his mother and the disciple whom he loved standing there.
The text never explicitly says that beloved disciple is male, nor that the beloved disciple is not one of the women just named.
Grammatically, Jesus could be entrusting his mother to another woman standing there.

John 21:24 distinguishes ``he who testifies'' (the beloved disciple) from the ``we'' who vouch for his testimony.
If later male leaders were uncomfortable with a woman as primary authority, you would expect: her role as beloved witness preserved, her name dropped, a masculine plural ``we'' to certify her testimony.
John 21:24 does exactly that.

\section{Women as Primary Witnesses and Narrative Protagonists}\label{sec:women-primary-witnesses}

\subsection{Mary Magdalene First in Lists}\label{subsec:mary-first-in-lists}

In lists of women in the Synoptics, Mary Magdalene is almost always named first, just as Peter is in lists of the Twelve.
Scholars like Carla Ricci note this structural parallel: Mary Magdalene's position among female disciples is analogous to Peter's among male apostles.

John gives the Beloved Disciple a role parallel to Peter's, but more faithful, more insightful, closer to Jesus.
Extra-canonical texts say Mary was loved more than the others, and that Peter resented it.
The simplest synthesis: the Johannine beloved disciple tradition is the masculinized, anonymized form of a memory that originally named Mary Magdalene as the disciple whom Jesus loved more than the others.

\subsection{Women at Narrative Hinges: Quantifiable Structural Pattern}\label{subsec:the-gospel-of-john-passes-the-bechdel-test.}

The strict modern Bechdel test (two named women speaking to each other about something other than a man) is a contemporary heuristic, not a tool of ancient literary analysis, and is not directly applicable to ancient theological narrative.
The Gospel of John does not preserve extended woman-to-woman dialogue on the page.
But if the test is adapted in a historically appropriate way---asking whether women speak in their own voice about God, truth, and the world rather than merely reinforcing male characters---John passes decisively.

The structural evidence is countable.
Across the Gospel's major narrative hinges---moments where the plot changes direction---women stand at every transition:

\begin{enumerate}
\item \textbf{Cana (John 2:1--11):} Jesus' first sign and the beginning of his public manifestation occur at the initiative of his mother.

\item \textbf{Samaritan woman (John 4:4--42):} the first extended theological dialogue and the first mass belief outside Judea arise from a woman's conversation and testimony.

\item \textbf{Martha (John 11:17--27):} the first explicit Christological confession (``I believe that you are the Christ, the Son of God'') is spoken by a woman.

\item \textbf{Mary of Bethany (John 12:1--8):} the transition into the Passion is marked by a woman's anointing, interpreted by Jesus as preparation for burial.

\item \textbf{Mary Magdalene (John 20:11--18):} the first resurrection encounter and the first apostolic commission (``go to my brothers'') are entrusted to a woman.
\end{enumerate}

These are five pivotal narrative inflection points of the Gospel.
Male disciples do have significant moments: the calling of the first disciples (John 1:35--51), the Bread of Life discourse (John 6), and Thomas's confession (John 20:28) are undeniably important.
But these function within narrative phases, not as turning points between phases.
When the story shifts direction---from private life to public ministry, from Judea to Samaria, from life to death, and from death to resurrection---the narrative agent at each threshold is a woman.
All five pivotal hinges are mediated by women's initiative or women's voices; none of the male disciples occupy any comparable structural position.

This distribution is highly unusual in the context of ancient Mediterranean narrative, where women typically appear in supporting, decorative, or cautionary roles.
John instead centers women as theological interlocutors, initiators of events, and primary witnesses of revelation.
The pattern is measurable, consistent, and deliberate.
It aligns closely with the hypothesis that the Gospel was composed by a woman whose perspective, memory, and priorities shaped the very architecture of the narrative.

\subsection{Note the women traveling with Jesus and the disciples are also linked to the royal family of Herod.}\label{subsec:note-the-women-traveling-with-jesus-and-the-disciples-are-also-linked-to-the-royal-family-of-herod.}

``The Twelve were with him, and also some women who had been cured of evil spirits and diseases: Mary (called Magdalene) from whom seven demons had come out; Joanna the wife of Chuza, the manager of Herod's household; Susanna; and many others.
These women were helping to support them out of their own means.'' Mary Magdalene, Joanna, and Susanna are specifically named as financial supporters.
Joanna was connected to Herod's court, meaning she had access to wealth and influence.
The phrase ``out of their own means'' implies they personally funded Jesus' ministry.
Multiple sites named Migdal existed in first-century Galilee, with at least two plausible candidates not far from Nazareth.
The text does not place Mary Magdalene in Herod's court; only Joanna is explicitly court-linked.
But if Mary Magdalene moved in similar elite Galilean circles, it would make sense that she would have been a skilled writer, yet perhaps not as skilled as the most prominent classics.
This explains the very deep knowledge of royal greek cult, while still maintaining some lack of literary sophistication.

\subsection{Male Pronouns Do Not Invalidate Female Authorship}\label{subsec:the-authorship-of-mary-is-typically-dismissed-as-a-possibility-because-of-the-use-of-male-pronouns-in-the-gospel.}

The standard objection claims that the Beloved Disciple must be male because the Gospel uses masculine grammatical forms: ὁ μαθητής (``the disciple'') and the masculine relative pronoun ὃν (``whom'').
This objection only works for people who think in English.
English has no grammatical gender.
English speakers therefore subconsciously assume that gender-marking in other languages must reflect biological sex.

Speakers of gendered Indo-European languages instantly know this is false.

In Polish, for example: Ofiara była mężczyzną (``The victim was a man'') and Ewangelista był kobietą (``The evangelist was a woman'').
Both are perfectly normal.
Ofiara is grammatically feminine; ewangelista is grammatically masculine.
The verb forms była and był follow the grammatical gender of the noun, not the sex of the person.
Every Czech speaker, German speaker, French speaker, Spanish speaker, Italian speaker, Lithuanian speaker, and Greek speaker understands this intuitively: grammar follows the noun, not the body.

This phenomenon is everywhere:
German die Person war ein Mann---feminine Person, masculine referent.
French la victime était un homme---feminine victime, masculine referent.
Spanish la persona era un hombre---feminine persona, masculine referent.
Arabic feminine al-daḥiyya (``victim'') used for men: kānat al-daḥiyya rajulan (``the victim was a man'').

English speakers struggle with this because their language does not encode grammatical gender at all, so they falsely import biological assumptions into Greek grammar.

Koine Greek---the common Greek that became the lingua franca of the eastern Mediterranean after Alexander's conquests---provides perfect parallels.
In papyri and inscriptions: μαῖα (``midwife,'' exclusively feminine form) is used for men (P.Oxy.\ 119, 744); ἡ γραμματεύς (``secretary,'' with feminine article) is used for a male official (P.Tebt.\ 51; P.Mich.\ 462).
Inside the New Testament, Phoebe in Romans 16:1 is διάκονος, a grammatically masculine noun, even though she is a woman---the participle οὖσαν marks female sex, the noun stays masculine.
Conversely, grammatically feminine nouns like ψυχή (``soul, person'') frequently refer to men.

The strongest New Testament parallel is 1 Peter 3:20, which explicitly speaks about Noah and his family:

\begin{quote}
ὅτε ἀπεξεδέχετο ἡ τοῦ Θεοῦ μακροθυμία ἐν ἡμέραις Νῶε, κατασκευαζομένης κιβωτοῦ, εἰς ἣν \ldots{} ὀκτὼ ψυχαὶ διεσώθησαν δι' ὕδατος.

``when God's patience waited in the days of Noah, while the ark was being prepared, in which \ldots{} eight souls were saved through water.''
\end{quote}

The structure is clear.
Νῶε (Noah) is explicitly named.
The phrase ὀκτὼ ψυχαί (``eight souls'') refers directly to Noah and his household---Noah plus seven others.
ψυχή is a grammatically feminine noun, and therefore the relative pronoun ἣν must also be feminine.
Noah is grammatically feminine in this sentence, not because Noah is female, but because ψυχή is a feminine noun.

This is an unambiguous internal New Testament demonstration that Greek grammatical gender follows the noun, not the sex of the referent.

This is exactly the same as Polish: Ofiara była mężczyzną (feminine grammar, male referent) and Ewangelista był kobietą (masculine grammar, female referent).

Once this is understood, the objection evaporates.
If Greek can refer to Noah with feminine morphology because ψυχή is feminine, then Greek can refer to the Beloved Disciple with masculine morphology because μαθητής is masculine.
Grammatical gender is not biological gender.
The masculine grammar in John cannot be used as evidence against female authorship.
It reflects the declension of μαθητής, not the gender of the author or the disciple.

\section{Transmission, Erasure, and Synthesis}\label{sec:transmission-erasure-synthesis}

\subsection{Textual Evidence of Women's Erasure}\label{subsec:textual-evidence-of-womens-erasure}

The hypothesis that a female author's identity was obscured is not speculation.
We have hard manuscript evidence that scribes deliberately modified female prominence in early Christian texts.

The clearest case is Junia in Romans 16:7.
Paul greets Ἰουνίαν (Junia) as ``outstanding among the apostles.''
The name is grammatically and historically female; no attested male form ``Junias'' exists.
Yet for centuries scribes and translators made it male, because they assumed no female apostle was possible.
Modern critical work by Eldon Epp, Michael Bird, and others has definitively shown this masculinization was ideological distortion.
This is a documented instance where a woman's apostolic status was obscured by later transmission.

The same mechanism is visible within the Gospel of John itself.
Elizabeth Schrader Polczer's article ``Was Martha of Bethany Added to the Fourth Gospel?'' (JBL 2017) demonstrates significant textual instability in the Mary/Martha passages of John 11--12.
The data are striking: early manuscripts including P66 and P75 show the names ``Mary'' and ``Martha'' shifting positions, with ``Mary'' appearing where later manuscripts read ``Martha,'' or the two being swapped entirely.
Some readings suggest that only one sister was originally present, named Mary, not Martha.
Most significantly, the climactic confession in John 11:27---``Yes, Lord, I believe that you are the Christ, the Son of God, the one coming into the world''---may originally have been placed in the mouth of Mary, not Martha.

Schrader Polczer's conclusion is straightforward: the textual evidence is consistent with the hypothesis that an original single strong female confessor named Mary in John 11 was later split into two figures and partially toned down, probably to reduce the prominence of that Mary.

This is not theory.
These are variant readings, name substitutions, and manuscript instabilities that attest deliberate scribal reshaping of female figures in John.
If the Johannine tradition once preserved an even stronger female Christological witness that scribes later tried to dilute, that is exactly the kind of process that could also turn ``Mary, beloved disciple'' into an anonymous ``disciple whom Jesus loved.''

\subsection{Mary Magdalene as the Fourth Pillar: Synthesis of the Evidence}\label{subsec:mary-magdalene-as-the-fourth-pillar-synthesis}

Bringing together the strands of manuscript evidence, narrative focus, early church controversies, and linguistic analysis, it becomes clear that Mary Magdalene was not merely a peripheral figure, but stood alongside Peter, James, and Paul as a foundational pillar of the early church. Her eyewitness testimony, preserved in the Gospel of John, shaped the community's understanding of Jesus' identity and resurrection. Recognizing Mary Magdalene as the Beloved Disciple elevates her to her rightful place as the ``apostle to the apostles,'' a status long obscured by later attribution and redaction. Alongside Peter, James, and Paul, Mary Magdalene emerges as the fourth great pillar of Christian origins.

This is not revisionist speculation.
Eastern Orthodox and Byzantine traditions preserve exactly this memory.
In the Greek and Slavic churches, Mary Magdalene is consistently titled ἰσαπόστολος---``Equal to the Apostles''---a status shared by only a handful of saints.
She is remembered as ``Myrrh-bearer and disciple of Christ,'' the one who followed him ``even to Golgotha'' and ``bore the news of his resurrection to the disciples.''
Annual liturgical hymns describe her as ``the one beloved of the Master, who stood with Him at the cross.''

There is also a longstanding Eastern tradition that Mary Magdalene traveled to Ephesus with the Apostle John and died there, her relics later being translated to Constantinople.
This is precisely the geographical orbit of the Fourth Gospel.
The memory that John's circle in Asia Minor was shaped by Mary's presence survives in Greek and Byzantine sources---outside the Latin West, where she was later reduced to a penitent ex-prostitute.

No early tradition explicitly identifies John son of Zebedee as ``the disciple whom Jesus loved.''
The earliest living memory of a ``beloved disciple'' is attached to Mary, not to John Zebedee.
Even if legendary in detail, the consistent Eastern memory of Mary as ``Equal to the Apostles,'' traveling with John and dying in Ephesus, looks like a later folkloric crystallization of an earlier historical reality: that Mary Magdalene's testimony was foundational in the Johannine community.

\subsection{Authorship Evidence Matrix: Mary Magdalene vs. Joanna}\label{subsec:authorship-evidence-matrix}

\begin{center}
\begin{tabular}{|l|c|c|}
\hline
\textbf{Authorship Evidence} & \textbf{Mary Magdalene} & \textbf{Joanna} \\
\hline
Present at Cross and Tomb & \checkmark & \checkmark (at tomb) \\
Named as Eyewitness in John & \checkmark &  \\
Central in Resurrection Narrative & \checkmark &  \\
Possible Link to Elite Circles & \checkmark & \checkmark \\
Early Church Rivalries (Peter vs.) & \checkmark &  \\
Tradition as “Apostle to Apostles” & \checkmark &  \\
Possible Redactional Role &  & \checkmark \\
Plausible Conduit for Traditions &  & \checkmark \\
\hline
\end{tabular}
\end{center}

\subsection{The Beloved Disciple is thought to have played an important role in the early Christian community.}\label{subsec:the-beloved-disciple-is-thought-to-have-played-an-important-role-in-the-early-christian-community.}

Some scholars argue that John, as a male apostle, would have had more influence in the early church, and thus, it seems unlikely that a woman like Mary Magdalene would have been the one to write or serve as a central figure.
The early church's view of Mary Magdalene is much more complex than sometimes assumed.
Though she was initially marginalized, she held an important role in early Christian traditions.
The Gospel of Mary, a non-canonical text, suggests that Mary Magdalene had an influential role, potentially even as a leader among the disciples.
Furthermore, early Christian communities were not as rigid in terms of gender roles as sometimes portrayed, and there is evidence that women held significant positions in the early church (e.g., Priscilla, Phoebe, Thecla).

\subsection{The Gospel of John emphasizes the Jesus was the only-begotten son of God.}\label{subsec:the-gospel-of-john-emphasizes-the-jesus-was-the-only-begotten-son-of-god.}

Unique Relationship with God: Jesus is depicted as having a unique and intimate relationship with God the Father, one that is unlike any other.
In John 1:14, 18, and 3:16, Jesus is referred to as the ``only-begotten Son'' (Greek: μονογενής, monogenēs).
This term emphasizes the uniqueness of Jesus as the Son, as no other figure shares this same divine sonship.

\subsection{Here there is a very interesting perspective linking the gnostic gospel of Thomas with John:}\label{subsec:here-there-is-a-very-interesting-perspective-linking-the-gnostic-gospel-of-thomas-with-john}

``Jesus said, `If they say to you, \texttt{Where\ did\ you\ come\ from?} say to them, \texttt{We\ came\ from\ the\ light,\ the\ place\ where\ the\ light\ came\ into\ being\ by\ itself,\ established\ itself\ and\ became\ manifest\ through\ their\ image.}If they say to you, \texttt{Who\ are\ you?} say, \texttt{We\ are\ its\ children,\ and\ we\ are\ the\ elect\ of\ the\ living\ Father.}'\,'' There are a few places where Thomas seems to indicate ``we'' are all children of God.
The uniqueness of the sonship seems to be very heavily emphasized in John to the point where it almost seems to become the core concept of the gospel.
Interestingly simultaneously John is making Thomas into the non-conformist and disbeliever.
ch 11.
All the disciples are afraid to go to Jerusalem, but Thomas says, ``Let us also go, that we may die with him.'' ch 14.
Jesus mentioning he will go to his father.
Thomas doubts.
Then Jesus says, ``I am the way, the truth, and the life.
No one comes to the Father except through me.'' You can only come to the father through the only-begotten son.
ch 20.
Thomas recognizes.
You are my Lord and my God.
This shows Thomas despite being Jesus brother or twin, likely holding a meaningful claim himself, John may try to emphasize Thomas could have contested Jesus at first, but then recognized him as the only rightful heir.

14 John 12:31 (NIV): ``Now is the time for judgment on this world; now the prince (archon) of this world will be driven out.'' In this verse, ``archon'' refers to Rome or the prince of this world.
It signifies authority and power over the world, but it's used in a negative sense to describe the ruler of evil.

\subsection{Temple destruction is typically used as a marker for the dating of the gospels.}\label{subsec:temple-destruction-is-typically-used-as-a-marker-for-the-dating-of-the-gospels.}

This argument is usually applied to the synoptic gospels, but needs a bit of a different treatment for John.
One note though that in case of John ``Destroy this temple, and I will raise it again in three days.'' Unlike in Olivet discourse, the identity and the prophecy of the destruction of the particular temple in Jerusalem in 70AD is much more questionable.
And that if we do not overemphasize the Mosaic beliefs of the gospel of John, then the temple could be one of the many many many other temples destroyed, not necessarily the main temple in Jerusalem.
Note that the temple in Jerusalem was already destroyed by the Babylonians and then rebuilt and already besieged by romans in 63BC.
It must not have been a very big stretch it could be destroyed again.
Sepphoris was destroyed by the Roman army around 4 BCE and rebuilt during Jesus' youth; as a τέκτων in nearby Nazareth, Jesus likely lived near this major reconstruction activity.

\subsection{The females of the greek period were very empowered but not in roman and jewish culture.}\label{subsec:the-females-of-the-greek-period-were-very-empowered-but-not-in-roman-and-jewish-culture.}

True---there were multiple female rulers in both the Ptolemaic dynasty and the Hasmonean kingdom.
The Ptolemies had several Cleopatras, Berenices, and Arsinoës ruling independently or as co-rulers.
Even in the Hasmonean period, Salome Alexandra (76--67 BCE) ruled as queen and was one of the most powerful figures of her time.
This definitely weakens the argument that a woman couldn't have written John just because of gender.
If elite women in those circles had power, wealth, and influence, then a female gospel author (like Mary Magdalene or another court-connected woman) wouldn't be impossible.
In John 11:27, Martha says: ``ναί, κύριε· ἐγὼ πεπίστευκα ὅτι σὺ εἶ ὁ Χριστὸς ὁ υἱὸς τοῦ θεοῦ ὁ εἰς τὸν κόσμον ἐρχόμενος.'' She uses ``Χριστός'' (Christos), not ``Μεσσίας'' (Messiah).
This is significant because John does use ``Μεσσίας'' in other places (John 1:41 and John 4:25), but here, Martha is speaking in Greek terminology.
This aligns with John's overall Greek imperial framework, rather than a purely Jewish messianic expectation.

John never uses Μεσσίας (Messiah) alone---he immediately translates it as Χριστός, meaning his audience wouldn't naturally recognize ``Messiah'' as a meaningful title.
John 1:41---``He first finds his own brother Simon and says to him, `We have found the Messiah' (which is translated Christos).'' (εὑρήκαμεν τὸν Μεσσίαν, ὅ ἐστιν μεθερμηνευόμενον Χριστός).
John 4:25---``The woman said to him, `I know that Messiah is coming' (who is called Christos).'' (Οἶδα ὅτι Μεσσίας ἔρχεται, ὁ λεγόμενος Χριστός).
Both times, it appears in direct speech, meaning these are Jewish characters (Andrew and the Samaritan woman) using a Jewish term, not the narrator or author.
This is the case where the greeks tried to convince the Jews to accept Jesus as their rightful ruler as well to additionally bolster his claim and gather more support.

Nathanael's declaration: ``Rabbi, you are the Son of God!
You are the King of Israel!'' This means Nathanael is recognizing that Jesus, who is acknowledged as Christos (a Greek royal title), is not only the ruler of the Greek world (Christos) but also the king of the Jews, in a more localized context.
The title ``King of Israel'' acknowledges the Jewish concept of kingship, which would have been understood as a ruler over Israel, distinct from the broader Greek context of kingship.
Andrew's declaration to Simon: ``We have found the Messiah.'' Andrew is indicating that Jesus, who is referred to as Christos (the Greek royal title), is not just the Greek king (Christos) but also the Messiah, a title that resonates specifically with Jewish expectations of a king.
He's telling Simon that Jesus fulfills both roles---the Greek Christos and the Jewish Messiah---and both Jewish and Greek audiences should recognize him as the ruler of both realms.

\subsection{In that time, kings and emperors were often called divine, not necessarily because people believed they were literal gods, but because divinity was part of royal language and legitimacy.}\label{subsec:in-that-time-kings-and-emperors-were-often-called-divine-not-necessarily-because-people-believed-they-were-literal-gods-but-because-divinity-was-part-of-royal-language-and-legitimacy.}

For Mary Magdalene, calling Jesus her ``God'' wouldn't necessarily mean she thought he was some abstract deity but rather that he was her true, anointed ruler---her Christos---just as emperors were called divine.
She saw him as the rightful king, the one destined to rule, and her personal devotion to him would naturally be expressed in the highest possible terms.
So, when she calls him ``Lord'' or when Thomas says ``My Lord and my God'' in John, it could reflect the same kind of royal language used for emperors---deep respect, loyalty, and recognition of his ultimate authority---rather than a fully developed theological doctrine of divinity.

\subsection{Watching him die would have been unbearable.}\label{subsec:watching-him-die-would-have-been-unbearable.}

She had placed all her faith, love, and devotion in him---not just as a ruler, but as the center of her world.
When the Romans crucified him, it wasn't just the death of a man; it was the destruction of everything she believed in.
One reading of John 20's affective narration permits the hypothesis that grief, searching, and recognition language carries the weight of lived experience rather than functioning only as doctrinal report.
Faced with such loss, visions, dreams, or an overwhelming sense that he couldn't really be gone are psychologically plausible---not as explanation of origins, but as the texture of memory that John's narrative preserves.
At that moment, she wasn't just an eyewitness; she was the one who couldn't let go.

\subsection{The resurrection story in John isn't just a theological claim---it's a deeply personal, emotional experience that reflects the profound impact Jesus had on her life.}\label{subsec:the-resurrection-story-in-john-isnt-just-a-theological-claimits-a-deeply-personal-emotional-experience-that-reflects-the-profound-impact-jesus-had-on-her-life.}

This explains why the resurrection account in John is so intimate and personal, focusing on Mary's grief, confusion, and joy.
This explains why it is only Mary who sees Jesus first, and why she doesn't recognize him at first---her vision is clouded by tears and despair.

\subsection{Description of male nudity}\label{subsec:description-of-male-nudity}

There is an unusual detail of Simon Peter being naked in the presence of the disciple whom Jesus loved.
This detail seems to have no place in the narrative, and it is not clear why it is included.

Although more sophisticated readings of this passage have been proposed, the most direct explanation stands out.
A woman, emotionally close to Peter or Jesus, remembering this odd, slightly intimate moment: ``He was basically naked, saw Jesus, and scrambled to cover up before jumping in.'' Peter being ``γυμνός''---which can mean fully naked or simply stripped for work, without outer garment---is a kind of detail a woman would be more likely to remember vividly and include in her account, especially if she was reminiscing the stories of her youth.
Feels less like theology, more like memory.

And then finally the Jesus's resurrection is described by the shroud being left in the tomb with the implication that Jesus was naked when he met Mary Magdalene.
Portraying Jesus as naked meeting a woman would be a very fit conclusion for a female author romance writer, but would make very little sense in any other scenario.

\subsection{The Argument from Bodily Silence: Zero Physical Descriptions}\label{subsec:the-lack-of-description-of-physical-appearance-of-jesus-in-john-seems-to-contradict-the-idea-that-the-author-was-an-eyewitness-a-woman-and-a-lover-of-jesus.}

John largely avoids portrait-style physical description---the kind that would tell you what someone looked like.
We learn functional details (``blind from birth,'' ``seamless garment''), but no one in the Gospel receives a portrait.
Not Jesus.
Not the Beloved Disciple.
Not Peter.
Not Mary Magdalene.
Not Nicodemus.
Not the Samaritan woman.
Not Lazarus.
Not Pilate.
Not Judas.
Not John the Baptist.

There is not one beauty adjective, not one bodily characteristic that would let you picture anyone.
Not ``handsome.''
Not ``young.''
Not ``strong.''
Not ``tall.''
Not ``radiant.''
Not ``fair.''
Not ``well-shaped.''
Nothing.

This silence is incompatible with ancient homoerotic male authorship.
In every ancient text authored by or coded by male homoerotic desire, you find: beauty language (κάλλος), central in gay Greek literature (Plato, Xenophon, Anacreon, Theognis); gaze language (``I looked,'' ``I desired,'' ``I beheld his form''); youth language (``the youth,'' ``the boy,'' ``beautiful youth''); idealized bodies; eroticized touch.
John has none of this.

The Beloved Disciple reclining ``in the bosom'' of Jesus is not eroticized, not described physically, given no bodily detail at all.
The scene is theologically, not physically, coded.
Peter's nakedness (γυμνός) in John 21 is comic, embarrassing, hurried---with no physical detail, no aesthetic appreciation, no erotic subtext.
It reads like a woman remembering a man who embarrassed himself, not a gay narrator encountering a naked male body.

This silence is, however, precisely what female authorship would predict.
Female authors and female memory keepers in antiquity frequently avoid describing men's bodies---not because they cannot, but because of modesty conventions, relational rather than visual memory, focus on emotion rather than form, and domestic rather than erotic attention.
They describe sensory, domestic, emotional detail instead: smell of perfume, smell of decay, tears, homes, gardens, private rooms.
This is exactly what John does.

The decisive clue is Jesus himself.
If any ancient author---especially one emotionally bonded to another man---were describing the physical Jesus they loved, they would eventually say something.
No height.
No face.
No beard.
No eyes.
No hair.
No complexion.
No beauty.
No youthfulness.
No strength.
No scars.
Nothing.

A homosexual male author---especially one allegedly in love with Jesus---would never write a biography of his beloved without any physical description.
A woman recounting memories of the one she loved, however, may deliberately avoid describing him in visual terms, yet describe the emotional and relational reality, using his voice, his presence, his intimacy, but not his body.
This is precisely the pattern in female grief literature, widow lamentations, women's personal letters in antiquity, and female-authored hagiography.
Women consistently describe what a person did or meant, not what that person looked like.

One partial explanation is that Philo of Alexandria, from whom the author of John seems to have borrowed many ideas, also avoided describing physical appearance.
But this only strengthens the female authorship case: a gay male author would simultaneously adopt Philonic theology and retain male aesthetic gaze.
John retains the first and completely lacks the second.

John reads like a woman's memory of Jesus, not a man's erotic admiration of him.

\subsection{John leaves out the institution of the Eucharist.}\label{subsec:john-leaves-out-the-institution-of-the-eucharist.}

The Last Supper is a key moment in the synoptic gospels, where Jesus breaks bread and shares wine with his disciples, establishing the Eucharist.
In John, the meal scene omits the words of institution and instead highlights the foot-washing (John 13:1--17), emphasizing servanthood and humility.
John's eucharistic theology appears elsewhere, especially in the Bread of Life discourse (John 6), but the institutional moment itself is absent from the Last Supper narrative.

In the context of Jesus being an observer of Greek religious traditions, the Eucharist would have been already a set tradition.
Unlike in Greek feasts, a Jewish passover meal would have been centered around the lamb, and bread and wine rituals would not have been the center of the meal.
Most notably the widespread around that time, Dionysian feasts would have been centered around the wine and bread rituals.
While drinking metaphorical blood and body of Dionysus would have been a widespread practice, the drinking of the blood and body in a Jewish banquet would have been considered a major blasphemy.

Even if we discount synoptic gospels and the letter to Corinthians, the Didache, a very early Christian text, also mentions the Eucharist as a well-established tradition.

If the Eucharist was a well-established tradition, and already a centerpiece of the Christian faith by the late first century, then how can one possibly explain it just slipped John's mind?

Also noteworthy, John describes Jesus talking about flesh and blood in the context of logos, so a modernized version of the Greco--Roman feast.
This is exactly a modern perspective Philo would have embraced, and it is very likely that the author of John was trying to convey the same message.

Justin Martyr accuses followers of Mithras of copying Christian rites, which he attributes to demonic imitation.
This polemical claim indicates that by the mid-second century, Christian Eucharistic practice was publicly recognizable enough to invite comparison with mystery-cult meals.

\subsection{John’s topographical accuracy confirms an early eyewitness.}\label{subsec:johns-topographical-accuracy-confirms-an-early-eyewitness.}

The Gospel of John describes specific places in Jerusalem such as the Pool of Bethesda with five porticoes, the Gabbatha pavement, and the Kidron valley.
These details were only rediscovered through modern archaeology, and would not have been available to a second-century writer relying on imagination.
This is a strong indication that John reflects either direct memory or testimony of someone present in the city before its destruction in 70AD.

\subsection{Semitic idioms preserved in John indicate early origin.}\label{subsec:semitic-idioms-preserved-in-john-indicate-early-origin.}

John’s Greek contains unmistakable Semitic phrases and even untranslated Aramaic words like ῥαββουνί (John 20:16), Μεσσίας (Messiah, immediately glossed as Christos in John 1:41 and 4:25), and Γολγοθᾶ (Golgotha, John 19:17).
These are not later church formulas but raw transliterations of speech, retained even when the author immediately translates them for a Greek audience.
Other turns of phrase, such as “Amen, amen, I say to you” (ἀμὴν ἀμὴν λέγω ὑμῖν), repeat the Semitic double-affirmation style, which is awkward in Greek but normal in Aramaic or Hebrew.
The phrase “living water” (ὕδωρ ζῶν, John 4:10) mirrors Semitic idioms (\texthebrew{יִם חַיִּים}, mayim ḥayyim) from the Hebrew Bible.
The density of these Semiticisms is too high to be explained as a Greek author “stylizing” his prose centuries later; they point instead to someone who thought in Aramaic or Hebrew and only secondarily expressed themselves in Greek.

This has profound implications for authorship.
If the Beloved Disciple was not one of the Twelve but a woman in Jesus’ circle, the preservation of such direct idiomatic phrases suggests her own voice or memory was carried into the final text.
Some early traditions associated Mary the mother of Jesus with reflections and sayings treasured in the Johannine community.
If she or another Palestinian woman shaped the source material, that would explain why John alone retains such stubbornly Semitic speech patterns, while the Synoptics smooth them out into more fluent Greek.
What we may be hearing in John, in other words, is not a stylized late theology, but the echo of real voices of Jesus’ earliest companions --- possibly even Holy Mary herself.

\subsection{Early liturgy reflects Johannine influence.}\label{subsec:early-liturgy-reflects-johannine-influence.}

The Didache, dated to the late first or very early second century, includes eucharistic prayers that resonate more with John’s themes of life, unity, and light than with the Synoptic Last Supper accounts.
This shows that Johannine theology was already shaping worship practices very early, undermining claims that John was a late innovation.

\subsection{Absence of explicit 70AD prophecy in John.}\label{subsec:absence-of-explicit-70ad-prophecy-in-john.}

If John had been written after the fall of Jerusalem, one would expect a triumphant prophecy-fulfillment motif.
Instead, John’s reference to the temple (“destroy this temple and I will raise it in three days”) is framed in symbolic, Christological terms rather than as a reference to 70AD.
The silence about the Roman siege is powerful evidence of an earlier horizon.

\subsection{Hostile witnesses accepted John’s authority early.}\label{subsec:hostile-witnesses-accepted-johns-authority-early.}

By the mid-2nd century, critics like Celsus and Jewish polemicists engaged Johannine themes without claiming the gospel was a recent invention.
If John had appeared late, its novelty would have been attacked.
Instead, its authority seems already entrenched, consistent with a text circulating decades earlier.

\subsection{Polycarp and Irenaeus preserve continuity.}\label{subsec:polycarp-and-irenaeus-preserve-continuity.}

Polycarp, active around 90AD, knew John’s community, and Irenaeus explicitly testifies to Johannine authorship.
The generational closeness makes it implausible that the gospel could have been fabricated late and still be universally accepted as apostolic.
The reception line itself supports early composition.

\subsection{Paul already had high Christology.}\label{subsec:paul-already-had-high-christology.}

Paul, writing in the 50s AD, describes Jesus in exalted terms: the agent of creation in 1 Corinthians 8, and the one given the name above every name in Philippians 2.
John is not inventing a new doctrine, but narratively framing what Paul already proclaimed.
Therefore, the old assumption that “high Christology means late date” is invalid.

\subsection{Philo of Alexandria already had high Christology.}\label{subsec:philo-of-alexandria-already-had-high-christology.}

It is widely acknowledged, though consistently understated, that the Fourth Gospel and the writings of Philo of Alexandria share the same theological air.
Philo, active in the first half of the first century, systematized a doctrine of the \emph{Logos} that already anticipated much of what modern scholarship calls Johannine Christology.
In his works the Logos is not a minor metaphor but the central mediator of divine sovereignty, described in a striking variety of roles:
the divine blueprint and instrument of creation (\emph{De Opificio Mundi}~20, 146),
the “first-born son” and “archangel” of God (\emph{De Confusione Linguarum}~146--147),
the cosmic high priest interceding for creation (\emph{De Specialibus Legibus}~1.81; 4.123--125),
the “image of God” and the “bond” holding the cosmos together (\emph{De Somniis}~1.215--239; \emph{De Cherubim}~35--36),
and the guide and teacher of souls on pilgrimage (\emph{De Migratione Abrahami}~6).
He consistently attributes to the Logos sonship, mediation, illumination, priesthood, and indwelling presence---exactly the categories John employs in his Prologue and throughout his Gospel.

The parallels are concrete and undeniable.
John 1:3 proclaims that “all things came into being through” the Logos, directly mirroring Philo’s teaching that the Logos is the instrument and blueprint of creation.
John’s description of the Logos as the “only-begotten Son” (1:14, 18) echoes Philo’s designation of the Logos as God’s “first-born son.”
The Gospel’s insistence that “no one has ever seen God … the Son has made him known” (1:18) restates Philo’s conviction that the Logos is the mediator bridging the invisible God and the visible world.
John’s dualism of light and darkness (1:4--5) parallels Philo’s Platonic language of intelligible light opposed to material ignorance.
When John declares that the Logos “tabernacled” among us (1:14), he is reproducing Philo’s claim that the Logos is God’s presence dwelling in the cosmos as in a temple.
Even the Prologue’s coupling of “grace and truth” (1:14, 17) recalls Philo’s repeated association of the Logos with God’s mercy and truth.
Later Johannine sayings (“I am the way, the truth, and the life,” 14:6) resonate with Philo’s image of the Logos as the guide of souls.
In short, virtually every key theme of the Johannine Prologue has a Philonic analogue.

Modern scholars have wrestled with how to characterize this relationship.
C.~H.~Dodd argued already that the Gospel of John draws heavily on the Logos-doctrine of Hellenistic Judaism, with Philo the clearest witness.
Harold Attridge memorably described John and Philo as “two riffs on one Logos”: they share the same conceptual framework, with John’s unique note being incarnation.
Peder Borgen pursued the parallels most fully, showing that John and Philo both read Exodus traditions---such as the manna and the tabernacle---through Logos-centered exegesis, making their overlap too specific to be coincidence.
Raymond E.~Brown, though more cautious, conceded that Philo’s Logos provided one strand of the background, alongside Jewish \emph{Memra} and Wisdom traditions.
David Runia has documented how later Christians freely drew on Philo, while stressing that direct dependence in John cannot be demonstrated beyond doubt.
Daniel Boyarin offered a Jewish-centered explanation: John’s Prologue is intelligible as a midrash on the “Memra” within Second-Temple Judaism’s binitarian monotheism, with or without Philo.
Yet this very point confirms Philo’s significance, since his writings show that such a “high Christology” was already part of Alexandrian Judaism.
Marian Hillar pressed this to its conclusion: Philo’s Logos-theology is itself already Christological, anticipating the categories John later applies to Jesus.

The cumulative effect is stark.
What John proclaims in the Prologue---that the divine Logos, long confessed as Son, Light, Mediator, and High Priest, has now taken flesh---is not a sudden innovation.
It is the crystallization of Alexandrian Jewish theology already articulated by Philo.
High Christology did not await the resurrection; it was alive in Greek Judaism before the lifetime of Jesus.

\subsection{John’s genre fits early Greco--Roman biography.}\label{subsec:johns-genre-fits-early-greco-roman-biography.}

The structure of John resembles Greco--Roman bios: carefully chosen public signs, extended speeches, climactic trial and death, and focus on legitimacy of rule.
This is consistent with first-century royal biography.
By contrast, second-century apocrypha adopt esoteric or sayings-only styles, which look later.
John therefore belongs in the earlier biographical stream.

\subsection{Summary and consequence.}\label{subsec:summary-and-consequence.}

Taken together --- the manuscripts P52 and P66, archaeological accuracy, Semitic idioms, liturgical echoes, silence on 70AD, hostile and friendly reception, continuity through Polycarp, consistency with Paul, bios genre, Synoptic texture, and apocryphal dependence --- all converge on a single conclusion.
The Gospel of John is not the latest, but one of the earliest.
The Synoptics too belong closer to the first generation than to the post-70 horizon.
The gospels function as royal biographies written quickly and spread swiftly, not theological fictions centuries later.
Most crucially, the Johannine eyewitness is best identified as Mary Magdalene---the fourth pillar alongside Peter, James, and Paul---with Joanna remaining a plausible conduit or transmitter, but not the primary witness.

\subsubsection{The Message}\label{subsubsec:the-message}

Finally, the traditional view of John being a man.
We have to keep in mind that most scholars agree the very early Christian communities had very strong influence from women.
Figures like Mary Magdalene, Tecla, and Priscilla were at the forefront of proclamation and patronage.
When the gospels were brought into Roman literary channels, the texts were reframed for a patriarchal audience.
Women leaders were erased from memory, Mary Magdalene was recoded as a prostitute, and post-Pauline voices were leveraged to silence women.
In that context a gospel authored by Mary Magdalene or Joanna would predictably circulate under the masculine title “John.”
The absence of a named “Beloved Disciple” in the Synoptics is therefore not a problem but a clue that the beloved witness is one of the women they do name.

What the Fourth Gospel calls its “signs” and “glory” functions as royal legitimation and as entry into \emph{secret teaching}.
John’s vocabulary of light and darkness, above and below, seeing and knowing, abiding and being born again, reads like a catechesis for an inner circle.
Jesus speaks privately to Nicodemus at night and to the Samaritan woman by a well, and he reserves the Farewell Discourses for the intimate circle, promising the Paraclete who will “teach you all things.”
This is precisely the pattern of \emph{mysteria}: public proclamation for all, and deeper instruction for those initiated.

Egypt is where that deeper instruction survived most visibly.
The Oxyrhynchus caches preserve Johannine fragments and cognate materials alongside documentary papyri that fix an early horizon.
The Nag Hammadi library near Chenoboskion preserves \emph{Gospel of Thomas}, \emph{Gospel of Philip}, \emph{Apocryphon of John}, and related tractates whose diction overlaps Johannine themes of light, truth, knowledge, and rebirth.
Mary Magdalene appears in the \emph{Gospel of Mary} as the authorized interpreter who out-reads Peter, exactly the rivalry implied by John’s Beloved Disciple scenes.
These Egyptian survivals were later classed as “heretical” under the broad label “Gnostic,” not because they were late fabrications, but because they preserved streams of \emph{secret teaching} that ran parallel to the increasingly Roman, clericalized canon.

Alexandrian intellectual culture makes sense of John’s Logos and its philosophical register.
Philo’s allegorical method and language of the Logos provide the city’s idiom in which Mary’s testimony could be framed for a Greek-reading audience.
The same city’s habit of school-based transmission explains why Johannine materials read like a handbook for advanced instruction rather than a simple chronicle.

Thus the message of John is double: royal proclamation and hidden wisdom.
It is the love story of Mary for her king, and the catechism of a court that trains emissaries.
What later bishops dismissed as “Gnosticism” is better understood as the \emph{esoteric tier} of the same royal gospel that the public tier proclaimed on the streets.
Read this way, attributing the gospel to “John” was not to deny the woman behind it, but to shield her voice so that her \emph{gnosis} could continue to feed the churches from Egypt to the ends of the Greek world.

\subsubsection{Miracles in John}\label{par:miracles-in-john}

There are seven miracles in John, which are often called the ``signs'' because they point to Jesus' identity and mission.
It is very worth pointing out that the miracles are not just random show of magic, but all are conveying the royal authority of Jesus.
Also, the miracles may not be completely a metaphor.
Jesus as a ruler very well learned could have been a skilled healer, or perceived as a skilled healer with the power of confirmation bias.
Notably it is not hard to believe that if Jesus visited a sick person and the person got better from just a placebo effect, then the person writing his royal biography would have been very keen to write about it.
Similarly, if Jesus had the resources to provide food for the hungry, and he showed the high generosity of a ruler, then the person writing his royal biography would have been very keen to write about it.

The seven signs in John are as follows:

\paragraph{1. Water into wine (John 2:1-11)} The transformation of water into wine signifies the arrival of a new covenant, just as royal feasts marked the legitimacy of kings.

\paragraph{2. Healing the official's son (John 4:46-54)} The healing of the official's son demonstrates his power over life and sickness, which in the ancient world were under divine royal prerogative.

\paragraph{3. Healing the paralytic (John 5:1-9)} The healing at the Pool of Bethesda demonstrates authority to restore what has been broken, a royal prerogative to make whole what society had abandoned.

\paragraph{4. Feeding the 5,000 (John 6:1-14)} Feeding the 5,000 mirrors the provision of a ruler to his people, akin to divine kingship in Hellenistic and Jewish traditions.

\paragraph{5. Walking on water (John 6:16-21)} Walking on water recalls dominion over the elements, a sign of imperial and divine control in both Greek and Jewish thought.

\paragraph{6. Healing the man born blind (John 9:1-12)} Granting sight to one blind from birth demonstrates power over congenital fate, the capacity to reverse what was thought irreversible from the moment of creation.

\paragraph{7. Raising Lazarus from the dead (John 11:1-44)} Healing the blind and raising Lazarus emphasize his authority over fate itself---challenging even the most immutable conditions of human existence.

\subsection{The non-miracle at the death of Jesus.}\label{subsec:the-non-miracle-at-the-death-of-jesus.}

It is also interesting to note that the gospel of John is the only gospel that does not mention the miracles at the death of Jesus.
As a gospel written by an eyewitness, it is very likely that the author was present at the crucifixion of Jesus, and would have been able to see the miracles that happened at the time.

\subsection{Conclusion}\label{subsec:conclusion}

The lack of mention of the miracles at the death of Jesus in the gospel of John is a strong indication that the author was actually attempting to write an accurate account of the events that happened at the time, and not just a theological treatise.
As if you were to introduce a miracle for theological reasons this would be the number one place to do it.
So although the seven miracles are not necessarily completely historically accurate, there is no compelling reason to believe they were not simply slightly exaggerated versions of the truth, the highlight of actual events where Jesus showed his leadership skills, knowledge of medicine, and generosity, just embellished as would have been expected from a royal biography, especially if the author was in love with Jesus.

What is really worth considering is that the events of the seven signs were really reported by the beloved disciple by Jesus as accurate.
For I tell you, let him who never exaggerated their manly deeds to their love interest be the one to cast the first stone.

\subsection{The Gospel of John point}\label{subsec:the-gospel-of-john-point}

While we discuss the historical Jesus and historical John, there is one thought that keeps coming back to me.

Nearly every discussion of the Gospel of John is about its theology.
We argue over doctrine and vocabulary and historic accuracy, but maybe we actually miss the single most important thing in the text.

We have lost the author.
We have lost the community that first carried it.
We have lost the Jesus they knew, and the beliefs of the first church.
What survives are copies of copies of copies, shaped by centuries of shifting theology and changing cultures.

But perhaps the best evidence for the true authorship of the Gospel of John is not the in any of the individual hard evidence, but rather the overall message of the Gospel itself.
What if the message of love is so strong, so deep, that it actually became the root of Christianity.
Perhaps when the billions of Christians read the Gospel of John, they may not realize, but they are subconsciously feeling the love that the author had for Jesus.
Perhaps an ancient letter about a lost love has become the most powerful and enduring message of love in human history.
Perhaps it is true that love is the only thing we're capable of perceiving that transcends the dimensions of time and space.

\section{Synoptic Gospels}\label{sec:synoptic-gospels}

\subsection{A minority reconstruction: Herod's death after 4BC and the apparent disagreement between Matthew and Luke on Jesus's birth date.}\label{subsec:herod-dies-in-10ad-not-4bc-and-the-gospels-of-matthew-and-luke-are-not-in-agreement-on-the-date-of-jesuss-birth.}

There is a small difficulty in dating the Jesus birth to 7BC, common date used in modern scholarship.

There is an apparent contradiction between the gospels of Matthew and Luke, as the gospel of Matthew states that Jesus was born during the reign of Herod the Great, who died in 4BC, while the gospel of Luke states that Jesus was born during a Roman census conducted under Quirinius, governor of Syria from 6 to 12 AD.
Roman censuses registered population and property for taxation; they were deeply resented in Judea because they signaled direct Roman control and triggered at least one armed revolt.

Even if we date the gospels very late to late second century, it is still hard to explain how the authors could be so right about so many historical background details, yet so wrong about such a critical part of the story.
Other than birth, Luke would have to be wrong about Jesus age at the time of baptism, which is also a very critical part of the story.
Adding to this, in this book we have a further discrepancy of Caspar, Melchior, and Balthazar journey in 7BC would make no sense as the Greek throne was still held by Strato III and this kingdom was not at the brink of collapse just yet.
Mainstream scholars dismiss contradictions between Matthew and Luke as proof of invention, but the eclipse correction to 12 AD shows both can be harmonized --- meaning both traditions may preserve genuine memory rather than fabrication.

It is very overstated what we really know historically about this date.
We know the Quirinius was governor of Syria from 6AD to 12AD .
It is worth emphasizing that Quirinius being a governor of Syria does not imply Judea was annexed the second day he came to power.
It does not violate any historical records to assume Judea was annexed just before he left the office in 12AD .
We know the census happened during his governorship, but we do not know when exactly.
Josephus does not claim the census occurred after the death of Herod the Great.
We know Josephus placed the death of Herod the Great at the lunar eclipse just before the Passover.
Note that he specifically emphasized he died after a long illness, which is consistent with the fact he may have spent a few years with the power shifted to his son Archelaus.

And here already comes a substantial mainstream scholarship error.
The passage from Josephus that is used to date the death of Herod the Great states ``the eclipse of the moon, which happened before his death, took place on the 10th day of the month Lous, which is Nisan.'' What is not admitted is that there were 51 lunar eclipses in the years 7BC to 12AD .
Out of these, there were three that happened at nighttime in Nisan and would have been prominent enough to be seen by the people.
The scholarly consensus is that this passage refers to the lunar eclipse of 4BC, however, upon closer inspection, in 4BC the lunar eclipse happened in Adar, a whole month earlier.

The only possible dates for the lunar eclipse that happened close to 10 Nisan between 4pm UTC and 4am UTC, so could have been seen by the people in Jerusalem is: 0012 Apr 24--22:38:24 10410 -24585 41 N -a 1.4693 0.1335 -0.8101 96.4 - - 11S 64E

Notably the lunar eclipse always happens at full moon and the 15 Nisan is the full moon of the month of Nisan.
However, as the lunar calendars do not align with solar, Jewish calendars introduce a second month of Adar, immediately before Nisan on leap years.
And it so happens that in year 12AD, the extra month of Adar moved the 15th of Nisan a few days later, which explains why Josephus would have said the lunar eclipse happened on the 10th of Nisan.
While on all other possible dates the lunar eclipse happened within a day of the 15th of Nisan.

This readjustment actually solves a lot of problems with the dating of Jesus's birth and the census.
And then the gospels would be in perfect agreement with each other.
Quirinius may have conducted his census in 6AD, Jesus could have been both in 6AD .
There was enough time to spend some time in Jerusalem in the first 2 years of Jesus's life, visiting the temple and being circumcised.
The magi could have come to visit him in 8AD when he was a very young child, right at the time of the fall of Greco--Indian kingdom.
Herod would have the time to order the slaughter of the innocents in 8AD, and Jesus could have fled to Egypt staying there for the next 4 years.
And then settling in Nazareth once Herod Antipas took over the throne of Judea.

With this new dating, there come some apparent contradictions with the records.
Most notably the traditional dating of the reign of Herod Archelaus, who was the son of Herod the Great and was the ruler of Judea from 4BC to 6AD .
Do note that Herod the Great was a tetrarch of Judea and not a king, and the title subsumed Judea proper Galilee Perea Idumea Samaria And other parts adjacent to Nabataea and the Decapolis.

At that time Archelaus was given the title of ethnarch, which was only a king of the Jews.
Josephus also quite clearly states that Archelaus ruled while Herod the Great was still alive.

\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Heir} & \textbf{Territory} & \textbf{Title} \\
\midrule
Archelaus & Judea, Samaria, Idumea & Ethnarch (not King) \\
Herod Antipas & Galilee, Perea & Tetrarch \\
Philip & Gaulanitis, Batanea, Trachonitis, Iturea & Tetrarch \\
\bottomrule
\end{tabular}
\end{center}

It is also important to highlight the tenures of the Roman governors of Judea is also actually tied to the reign of Herod Archelaus and the know dates are not attested from other sources.
So if the reign of Herod Archelaus ended in 12AD, there is still plenty of time to plausibly fit Coponius, Ambivulus, Annius Rufus and Valerius Gratus, the predecessors of Pontius Pilate, into the timeline.

\subsection{The main argument for the dating of the synoptic gospels to after year 70AD is the destruction of the temple by the Romans.}\label{subsec:the-main-argument-for-the-dating-of-the-synoptic-gospels-to-after-year-70ad-is-the-destruction-of-the-temple-by-the-romans.}

\subsection{Most scholars argue the Olivet Discourse in the synoptic gospels is a prophecy of the destruction of the temple.}\label{subsec:most-scholars-argue-the-olivet-discourse-in-the-synoptic-gospels-is-a-prophecy-of-the-destruction-of-the-temple.}

\subsection{Stylistic markers suggest a redactional layer in Mark 13.}\label{subsec:at-the-same-time-the-textual-analysis-of-mark-shows-that-these-verses-were-added-later.}

which is strikingly more formal and technical than the surrounding Markan language.
In the rest of the Gospel, Mark tends to favor more direct and accessible expressions of Jesus' teachings and prophecies, often focusing on clear, vivid imagery.
The ``abomination of desolation'' is a very specific, technical term that feels more aligned with prophetic, liturgical, or later theological writing.
This suggests it could be an editorial addition, introduced to reflect later theological concerns (like the destruction of the Temple in 70 CE) or to align the passage with more familiar Old Testament imagery.

\subsection{\texorpdfstring{``But in those days, after that tribulation\ldots{}'' (Mark 13:24)}{4.
``But in those days, after that tribulation\ldots'' (Mark 13:24)}}\label{subsec:but-in-those-days-after-that-tribulation-mark-1324}

The phrase ``after that tribulation'' introduces a more formal, structured apocalyptic expression that contrasts with the more immediate and direct tone of earlier sections in Mark.
Mark's Gospel typically uses simple, conversational language, but this phrase carries a more stylized apocalyptic tone, characteristic of later prophetic writings.
It introduces a period of cosmic upheaval, which is a prominent feature of later apocalyptic literature (including the Book of Revelation and some Jewish apocalyptic texts), but less common in the earlier parts of the Gospel.

\subsection{``When you see the desolating sacrilege set up where it ought not to be'' (Mark 13:14)}\label{subsec:when-you-see-the-desolating-sacrilege-set-up-where-it-ought-not-to-be-mark-1314}

Language Issue: This passage contains complex terminology like ``desolating sacrilege'' and the phrase ``set up where it ought not to be''.
While the ``abomination of desolation'' in earlier prophecies is a well-established term, this variation feels more convoluted and distant from the language of Mark's usual teachings.
It introduces an additional layer of complexity that seems more fitting to later apocalyptic writers, who used this terminology to indicate ritualistic desecration in a manner distinct from Mark's usual straightforward style.

\subsection{``Heaven and earth will pass away, but my words will not pass away'' (Mark 13:31)}\label{subsec:heaven-and-earth-will-pass-away-but-my-words-will-not-pass-away-mark-1331}

Language Issue: The eternal nature of Jesus' words is a theme more common in Johannine literature (John 1:1, 6:68, 14:23).
While Mark presents Jesus as authoritative, this statement emphasizes a long-lasting theological development that might be at odds with Mark's more human-centric portrayal of Jesus.
This phrase might also echo themes found in later Christian writings that focus on the eternal validity of Jesus' teachings in a way that is somewhat different from the earthly mission Mark emphasizes.

\subsection{``But concerning that day or that hour, no one knows, not even the angels in heaven, nor the Son, but only the Father'' (Mark 13:32)}\label{subsec:but-concerning-that-day-or-that-hour-no-one-knows-not-even-the-angels-in-heaven-nor-the-son-but-only-the-father-mark-1332}

Language Issue: The emphasis on the ignorance of the Son regarding the coming of the Kingdom contrasts with the high Christology that develops in later texts like John, where Jesus has full knowledge and authority.
While it is possible that this reflects Mark's emphasis on Jesus' humanity, the way the phrase is worded feels slightly out of sync with the rest of the Gospel, where Jesus is shown to have a strong authority over time and knowledge of God's plans.
The focus on ignorance is strikingly different from the confident tone found in later Christian reflections on Jesus.

\subsection{``Let the reader understand'' (Mark 13:14)}\label{subsec:let-the-reader-understand-mark-1314}

This phrase alone signals editorial activity.
Such asides appear when an author or redactor addresses the reader directly, clarifying a point that might otherwise be unclear.

\subsection{If Mark 13 and Mark 16 reflect later redactional expansion, when would these edits have been made?}\label{subsec:now-the-question-is-if-mark-13-and-mark-16-were-added-to-the-gospel-known-to-us-then-when-would-the-edits-have-been-made}

Well, likely it was quite soon after the event when the event was still fresh in the minds of the people.
As so many events relevant to global politics, Jewish and Christian history happened at the time, the edits were done say decades later, likely different events would have been on top of the minds of the editors to make the edits.
Based on this alone, we could put the upper end for the dating of the original manuscript of Mark to before 70AD.

\subsection{The other arguments for the lower bound of the dating of the gospels focus on the need to develop the theology of the early church.}\label{subsec:the-other-arguments-for-the-lower-bound-of-the-dating-of-the-gospels-focus-on-the-need-to-develop-the-theology-of-the-early-church.}

In the context of Jesus being an heir to the greek empire, the gospel of Mark would not require any development of the theology as it is a historical account of the final downfall of the greek empire.
The theological part of it would be the resurrection of Jesus, which was not present in the original gospel of Mark.
As such there is little reason to expect much time to pass between the historical events and the writing of the gospel of Mark.

\subsection{The gospel of Mark starts with the baptism of Jesus by John the Baptist.}\label{subsec:the-gospel-of-mark-starts-with-the-baptism-of-jesus-by-john-the-baptist.}

The baptism of Jesus is a very important event in the gospels, as it is the event that marks the beginning of Jesus' reign as the rightful king of the greek empire.
Courtly scribes often start at the beginning of the reign of the king, and the baptism of Jesus is likely the event that would have been considered the beginning of his reign.
Traditional scholarship does not provide a good explanation for why the story of Jesus would start with the baptism by John the Baptist, but in the context of Jesus being the rightful heir to the greek empire, it makes perfect sense.
The story of baptism is also portrayed as a very public event, which would have been a very important event for the people to know about.
In all four Gospels, during Jesus' baptism, a voice from heaven declares: ``This is my Son, whom I love; with him, I am well pleased'' (Matthew 3:17, Mark 1:11, Luke 3:22, John 1:34).
This is significant because, in ancient Jewish and Hellenistic traditions, the declaration of someone as a ``son of God'' often had royal connotations, signifying divine legitimacy or appointment to a kingship.
The title ``Son of God'' was used for kings in the Hebrew scriptures (e.g., 2 Samuel 7:14) and in Hellenistic traditions for emperors and rulers.

The baptism of Jesus follows the established pattern of royal coronation in the ancient Mediterranean world.
Psalm 2, the royal enthronement psalm of Israel, provides the governing framework: ``You are my son; today I have begotten you'' (Psalm 2:7).
This is the formula spoken by God to the newly installed king---not a statement about biological sonship, but a declaration of political legitimacy and divine appointment.
When the heavenly voice at Jesus' baptism declares ``This is my Son,'' it is invoking this same royal language.

The traditional reading of John the Baptist's ministry rests on the phrase ``repentance for the forgiveness of sins'' (μετάνοια εἰς ἄφεσιν ἁμαρτιῶν), which appears in Mark 1:4 and Luke 3:3.
But this translation obscures the lexical range of the Greek and imports later Christian moral theology into a first-century Judean context.
The Greek word μετάνοια does not primarily mean ``feel remorse about moral guilt''; it means ``change of mind'' or ``reconsideration,'' often used in political or strategic contexts to describe a reversal of stance or allegiance.
Similarly, ἄφεσις fundamentally means ``release'' or ``sending away,'' and in the Septuagint it is the term for Jubilee-style release of debts, slaves, and land obligations (Hebrew \texthebrew{דְּרוֹר} \textit{deror}, \texthebrew{שְׁמִטָּה} \textit{shemittah}).
The word ἁμαρτία, typically rendered ``sin,'' derives from a verb meaning ``to miss the mark'' or ``to err''---Aristotle's \textit{hamartia} in tragedy refers to a tragic error or misjudgment, not inherent moral evil.

Read in this lexical context, the phrase μετάνοια εἰς ἄφεσιν ἁμαρτιῶν can be understood as ``a change of allegiance for the release from prior obligations and failures''---language far more suited to a political or ritual transition than to private moral contrition.
This reading gains further support from the geographic and cultural setting: John operated in the Transjordan region, specifically Perea (cf. John 1:28, ``Bethany across the Jordan''), a Herodian territory east of the Jordan River heavily influenced by Greek administration and culture.
This liminal, Hellenized zone was a classic staging ground for alternative claimants and movements operating outside the direct control of Jerusalem's priestly establishment.

The text itself supports this reading.
In John 3:25, when a dispute arises about John's baptism, it is described as περὶ καθαρισμοῦ---``concerning purification.''
The Greek word καθαρισμός refers to purification, not μετάνοια (repentance) or ἄφεσις (forgiveness).
This is the language used for ritual preparation before taking sacred office, not for moral confession.
The Gospel of John categorizes the baptism as a consecration rite, not a repentance ritual.

Critically, the entire Gospel of John avoids the vocabulary of repentance altogether.
The words μετάνοια and μετανοέω (``to repent'') never appear in John's Gospel.
Instead, the Fourth Gospel consistently frames the movement in terms of purification (καθαρισμός), new birth (γεννηθῆναι ἄνωθεν), washing, and belief.
This is not an accident of style; it represents a fundamentally different conception of what John's baptism accomplishes.
Where the Synoptic Gospels use the language of repentance inherited from prophetic calls to covenant renewal, John uses the vocabulary of ritual transformation and identity change---exactly what one would expect if the baptism functioned as a royal investiture rather than moral reform.

Once the baptism is understood in this ritual category, the coronation structure becomes evident.
The baptism narratives contain a fixed three-part sequence found in ancient coronation ceremonies: ritual washing, divine anointing or bestowal of power, and public proclamation of royal sonship.
This pattern appears in the coronation traditions of Egypt, the Near East, and Israel, but not in repentance rites or prophetic commissioning.
The Gospel accounts present all three elements in order: Jesus undergoes immersion (washing), the Spirit descends upon him (divine anointing), and a voice from heaven proclaims ``This is my Son'' (royal proclamation).
Egyptian coronation rituals follow this same sequence---the king is purified, receives divine investiture from the gods, and is proclaimed the rightful ruler.
The same structure appears in Assyrian royal inscriptions, where kings are ritually prepared, receive divine favor, and are declared chosen by the gods.
David's anointing in 1 Samuel 16 follows the identical pattern: a ritual act by Samuel, the Spirit of the Lord rushing upon David, and his recognition as God's chosen king.

Ritual purification before assuming royal or priestly office was standard practice across the ancient Mediterranean and Near East.
The geographic context is crucial: John the Baptist operated in the Transjordan region (Bethany beyond the Jordan, John 1:28), which was under Greek political and cultural influence, not in Judea proper.
In this Hellenistic context, Greek purification rites (καθαρισμοί) were used for role transitions, prophetic commissioning, and rites before kingship---not for moral repentance in the later Christian sense.
In Ugaritic texts from Bronze Age Syria, kings performed lustration ceremonies---ritual washings---at designated times as part of their cultic and political duties, purifying themselves before offering sacrifices and performing royal functions.
The Hasmonean rulers of Judea, who combined the roles of king and high priest, were bound by priestly purity regulations that included mandatory ritual washing, prescribed vestments, and consecration before entering sacred space or performing cultic acts.
Greek and Roman traditions likewise employed purification rites (κάθαρσις in Greek, \textit{lustratio} in Latin) for cities, armies, magistrates, and sacred officials, using water, fumigation, and sacrifice to remove pollution (\textit{miasma}) before undertaking public functions.
The concept of ritual cleansing as preparation for a new political or sacred role was not peripheral but structurally central to ancient conceptions of legitimate authority.

What is crucial is that these purification rites were not about moral guilt but about status transition: they marked the passage from one condition to another, releasing the individual from previous obligations and preparing them for new responsibilities.
This is why ἄφεσις (``release'') is the appropriate term---not forgiveness of personal sins, but formal discharge from prior bonds so that a new office or identity can be assumed.

This coronation reading is further supported by the internal narrative logic, which contradicts the repentance interpretation.
In Matthew 3:14, John protests Jesus' baptism, saying ``I need to be baptized by you, and do you come to me?''
This protest makes no sense if the baptism is about moral repentance---John would simply perform the rite as he did for all others.
But if the ritual elevates the recipient to a status above the officiant, conferring royal authority, then John's reluctance is perfectly logical: he feels unworthy to perform a rite that places Jesus in a higher position than himself.
This is exactly what a coronation officiant would say when asked to anoint a superior.
The repentance model requires inventing later theological explanations for this protest; the coronation model explains it naturally within the narrative.

Early textual variants confirm this coronation reading by preserving the more explicit Psalm 2 formula.
The Gospel of the Hebrews and some manuscripts of Luke read ``Today I have begotten you'' at the baptism, quoting Psalm 2:7 directly rather than the softer ``This is my beloved Son.''
This is the exact wording of the Davidic enthronement formula---the declaration made at the moment a king takes the throne.
The presence of this variant shows that early Christian communities understood the baptism as Jesus' coronation, the moment he was formally installed as the Messiah-King.
Later manuscripts suppressed this reading when the political implications became problematic, just as references to the temple destruction and certain resurrection accounts were modified to emphasize religious rather than dynastic themes.

The introduction of Mark is best read as a coronation narrative rather than a generic opening to the ministry of Jesus.
When Mark writes, “The beginning of the good news about Jesus the Messiah, the Son of God,” he is not giving a neutral preface but invoking the precise \textit{euangelion} language of coronation, the same word used in imperial decrees such as the Priene inscription to proclaim the reign of Augustus.
When he cites Isaiah 40:3 --- “a voice of one calling in the wilderness, ‘Prepare the way for the Lord, make straight paths for him’” --- he is not simply quoting prophecy but announcing the royal entry of Jesus.
Mark then describes the gathering of the crowds: “The whole Judean countryside and all the people of Jerusalem went out to him,” which reads as the assembly of witnesses at an enthronement.
John follows with the formal announcement of the successor: “After me comes the one more powerful than I, the straps of whose sandals I am not worthy to stoop down and untie.”
Jesus then enters the scene, is baptized, and the heavens are opened.
The descent of the Spirit “like a dove” and the heavenly declaration, “You are my Son, whom I love; with you I am well pleased,” function as the acclamation of divine sonship.
Finally, Jesus withdraws into the wilderness for forty days.
This period of fasting and testing parallels the purification and trial rites known from Ptolemaic and Near Eastern traditions that prepared rulers for kingship.

This whole sequence is often divided into separate pericopes in modern editions, which obscures its unity.

When Mark writes that the Spirit descended "like a dove," this is not an arbitrary image.
Zeus was widely known as relaying his divine messages through a white dove.
Hesiod describes the doves of Dodona at one of the oldest and most famous oracles of Zeus: the priestesses were called "doves," and prophecy itself was said to be carried from the oak of Zeus.
In this context, the dove was a recognized emblem of Zeus' descent and communication, in the same way that the Holy Spirit later came to be portrayed in Christian tradition.

This event is often separated into separated titled chapters that confuse the reader into thinking these are separate events, not a continuous narrative of a single event.
Yet read continuously, this description of a coronation ceremony is unmistakable.
Every coronation in Ptolemaic Egypt was accompanied by a ritual of purification, fasting, and testing in the desert.
Zeus descended as a white dove to speak to the public is also an exceedingly common motif in Hellenistic royal iconography.

\subsection{11.3 Liturgical hymns as royal proclamation}\label{subsec:liturgical-hymns-as-royal-proclamation}

Another strong indication that no long interval of theological development was required before the composition of the gospels is the presence of complete hymnic material already embedded in them.
These hymns are not random prayers but are structured in the same idiom as Greek royal and imperial hymns of the Hellenistic and Roman worlds.
They read as coronation songs transferred directly from Augustus or the Ptolemies onto Jesus.

\paragraph{The Magnificat (Luke 1:46--55).}
Mary’s song is usually treated as a Jewish psalm echo, but its literary form is that of an enthronement hymn.
Its opening, “My soul magnifies the Lord, my spirit rejoices in God my Savior,” uses the verb \emph{megalyno} (“magnify”) and the title \emph{sōtēr} (“savior”), both key terms of Hellenistic ruler hymns.
Ptolemaic decrees and Augustan inscriptions hailed the monarch as “Savior of the world” and began with magnification formulas almost identical in tone.
The middle section describes a great reversal: “He has scattered the proud… put down the mighty from their thrones and exalted those of low degree… filled the hungry with good things and sent the rich away empty.”
This is not only Jewish prophetic rhetoric; it is the stock structure of imperial accession hymns, in which the new ruler humbles tyrants, exalts the poor, and distributes grain and wealth to his people.
Inscriptions from the reigns of Augustus and Nero praise the emperor precisely for “casting down the arrogant, lifting up the humble, and filling the hungry with bread.”
The closing lines (“He has helped his servant Israel in remembrance of his mercy”) parallel the endings of Hellenistic encomia that ground the ruler’s beneficence in ancestral tradition and divine mandate.
In sum, the Magnificat is a royal hymn: it declares that the child born of Mary is the true benefactor king who enacts the reversals celebrated in imperial propaganda.

\paragraph{The Christ hymn (Philippians 2:6--11).}
Even more striking is the pre-Pauline hymn quoted in Philippians:
“Though he was in the form of God, he did not grasp at equality with God,
but emptied himself, taking the form of a slave,
becoming obedient unto death---even death on a cross.
Therefore God highly exalted him
and gave him the name above every name,
that at the name of Jesus every knee should bow,
in heaven and on earth and under the earth,
and every tongue confess that Jesus Christ is Lord.”
The structure is unmistakable: descent in humility, vindication, exaltation, and universal acclamation.
This is exactly the grammar of Greek imperial hymns and acclamation formulae.
In the provinces, subjects were required to bend the knee and confess the lordship of Augustus or Tiberius; the same wording is preserved in dedicatory inscriptions that command “every nation and every tongue” to proclaim the emperor’s titles.
The giving of “the name above every name” matches the ceremonial bestowal of divine epithets on the ruler---\emph{Sōtēr}, \emph{Euergetēs}, \emph{Kyrios}.
The hymn ends in a universal chorus of submission, precisely the setting of an imperial enthronement liturgy.
Yet in this hymn the honorifics are not transferred to Caesar but to Christ, crucified and risen.
This is a direct appropriation of imperial forms for a rival claimant.

\paragraph{Conclusion.}
These two hymns alone are sufficient to overturn the common scholarly claim that theology required decades of fermentation after the crucifixion.
On the contrary, the gospels already contain imperial liturgical forms fully reworked around Jesus.
This is not slow doctrinal development; it is immediate political proclamation in the idiom of Greek hymnody.
The Magnificat enthrones the unborn child as benefactor king; the Philippians hymn acclaims the crucified as universal Lord.
Both show that from the very beginning the Christian message was cast in the established courtly language of the empire, declaring Jesus to be the rightful ruler.


\subsection{Gospel of the Hebrews variants of the baptism story}\label{subsec:gospel-of-the-hebrews-variants-of-the-baptism-story}

In some fragments, Jesus is portrayed as baptizing others, not merely receiving baptism.

This reverses the standard narrative and places Jesus in the position of the initiator, the one who commands and empowers.
That act reflects imperial privilege---a king is not merely initiated, he is the one who now administers the rites of his reign.
This underscores Jesus as king over a new order, already exercising regal authority.

The Gospel of the Hebrews uniquely calls the Holy Spirit Jesus's ``Mother,'' and she lifts him up by his hair and carries him to Mount Tabor.
This echoes ancient royal investiture scenes, where a deity (or female embodiment of Wisdom) reveals or blesses a new ruler.
Mount Tabor is symbolic --- a ``high place'' often used in imperial and divine appearances in both Jewish and Hellenistic traditions.
The motif of being lifted and carried by the Spirit recalls imperial enthronements where the new king is symbolically raised.
Although preserved only fragmentarily through patristic quotation, these traditions reflect a Jewish-Christian stream that retained royal-investiture and maternal-Spirit imagery absent from the later canonical synthesis.

\subsection{Royal Psalms:}\label{par:royal-psalms}

The Psalms contain numerous references to the king being God's anointed (e.g., Psalm 2:7, Psalm 45:7), and the baptism scene echoes these royal motifs.
By identifying Jesus as the Son of God and anointing him with the Spirit, the Gospel writers could be alluding to these royal traditions, suggesting a coronation-like event rather than merely a religious cleansing.

\subsection{The dove descending on Jesus at his baptism is a symbol of divine approval and empowerment.}\label{subsec:the-dove-descending-on-jesus-at-his-baptism-is-a-symbol-of-divine-approval-and-empowerment.}

The Dove as a Divine Symbol: In Greek and Hellenistic cultures, the dove was often associated with various divine manifestations, particularly linked to gods like Zeus and Apollo.
A dove descending from heaven was a common symbol of divine approval or empowerment, as in the case of Zeus transforming into a dove in some myths.
In Hellenistic royal ideology, the idea of divine signs, such as a dove appearing at the moment of a king's or leader's coronation, was not unusual.
Dove and the ``Anointing'' of Rulers: Greek rulers and Roman emperors were often seen as divinely chosen, and symbols of divine approval---such as a dove---would serve to emphasize this idea.
The use of a dove in the baptism of Jesus could be interpreted as a sign that Jesus is divinely anointed for kingship, in line with how Hellenistic rulers were often anointed by the gods.

\subsection{The gospel of Matthew includes the details of the birth of Jesus with the visit of the Magi.}\label{subsec:the-gospel-of-matthew-includes-the-details-of-the-birth-of-jesus-with-the-visit-of-the-magi.}

The visit of the Magi has been universally accepted as an allegory but as mentioned in the previous chapter, the titles of the Magi are exceedingly hard to fabricate by even the most knowledgeable of scholars in the Roman empire that were not greek empire loyalists.
The Herod's massacre of the innocents is also universally accepted as allegory, but as outlined in the Jesus Dynasty, the massacre of the innocents likely refers to Herod's killing of multiple of his own children and their families.
In the context that Joseph was likely too of the Hasmonean family, giving birth in the city of the Hasmonean dynasty, Bethlehem, is again a lot more likely to be a real event than an allegory.
Similarly, the flight to Egypt is likely a real event as a noble family would have had more than enough resources to flee to and be welcomed in Egypt for a few years.

\subsection{Finally, the gospel of Matthew includes the great commission, which is likely an edit to the original gospel of Mark.}\label{subsec:finally-the-gospel-of-matthew-includes-the-great-commission-which-is-likely-an-edit-to-the-original-gospel-of-mark.}

πορευθέντες οὖν μαθητεύσατε πάντα τὰ ἔθνη Make disciples of all nations---Spread the teachings of Jesus to people all over the world.
Here the word ``nations'' is a translation of the Greek word ``ethne,'' which can also be translated as ``Gentiles'' or ``non-Jews.'' Baptize them in the name of the Father, the Son, and the Holy Spirit---Initiate them into the faith through baptism.
Teach them to obey everything I have commanded you---Continue teaching and guiding new believers to follow Jesus' teachings.
Jesus promises his presence---``And surely I am with you always, to the very end of the age.'' The great commission is essentially what the emissaries of the greek empire would have been doing, spreading the decrees of the king across the empire.

\subsection{The term ἔθνη is typically translated as ``nations'' but can also be translated as ``Gentiles'' or ``non-Jews.''}\label{subsec:the-term-ux1f14ux3b8ux3bdux3b7-is-typically-translated-as-nations-but-can-also-be-translated-as-gentiles-or-non-jews.}

In the context of the greek empire, the term would have been used to refer to all the greek people across the empire, all the nations, kingdoms or tribes that were part of the greek empire.

\subsection{The temptation of Jesus in the wilderness following the baptism.}\label{subsec:the-temptation-of-jesus-in-the-wilderness-following-the-baptism.}

Immediately after the baptism---his anointing as Son and Christ---Jesus is led into the wilderness for forty days of fasting (Mark~1:12--13; Matt.~4:1--11; Luke~4:1--13).
There he faces a threefold royal ordeal: \emph{bread} (material provision), \emph{spectacle} (a miracle to prove divine favor), and \emph{kingdoms} (world sovereignty).
John preserves the same logic across the ministry rather than in one scene: crowds want bread (John~6:26), relatives push public signs (John~7:3--4), and the people attempt to make him king by force (John~6:15).
This is not presented as a prophet checking boxes of “fulfilled verses,” nor as a tale borrowed from one foreign school.
It reads as a coronation‐threshold trial drawn from a repertoire as old as kingship itself.

\paragraph*{Egyptian background (pharaonic accession).}
In Egyptian state theology the king’s ascent required ritual victory over \emph{isfet}---chaos, falsehood, the desert “red land”---in order to establish \emph{maʿat}---order, truth, justice.
Coronation cycles pair purification (abstinence, lustrations, vesting) with the liturgical “smiting of enemies,” divine speeches, and bestowal of regalia.
Examples of this polarity and accession rite pattern are visible already in the Pyramid and Coffin Texts, and in temple programs at Karnak, Luxor, and Medinet Habu, culminating in the Edfu cycle known as the “coronation of the falcon.”
The geography matters: in Egyptian imagination the wilderness belongs to chaos.
Mark’s note that Jesus was “with the wild beasts” situates his trial precisely in that zone of disorder.
Form differs (public rite vs.\ solitary fast), but function aligns (ordeal in the domain of chaos immediately before rule).

\paragraph*{Israelite background (Moses, Israel, David).}
Moses spends \emph{forty days} on Sinai “without bread or water” before receiving the Law (Exod.~34:28; Deut.~9:9).
Israel is said to have been “tested” \emph{forty years} in the wilderness by hunger, by “testing God,” and by loyalty to him alone (Deut.~8:2--3; 6:13, 6:16)---the very verses Jesus cites in reply to Satan.
David’s accession ordeal is ethical rather than ascetic: twice offered the shortcut of killing Saul, he refuses to seize the throne (1~Sam~24; 26).
These are not the same story as the Gospels, but they teach the same kingship rule: legitimate rule comes through obedience and ordeal, not grasping.

\paragraph*{Indo-Iranian and Greco-philosophic comparanda.}
The most \emph{precise} narrative parallel is Siddhartha Gautama’s trial under the Bodhi tree: Māra offers (1) sensual desire, (2) a terrifying challenge demanding miraculous proof of survival, and (3) sovereignty over the world, as related in the \emph{Lalitavistara} (ch.~24) and the \emph{Nidānakathā}.
The structural match to the Gospel’s triad of \emph{bread-spectacle-kingdoms} is striking.
Zarathustra’s confrontation with Angra Mainyu likewise unfolds as a triad: (1) riches and wealth, (2) women and pleasure, and (3) kingship and power; he refuses each in fidelity to \emph{asha} (truth/order), as preserved in the \emph{Vendīdād}~19 and the \emph{Dēnkard}~7.
Greek moral pedagogy offers the Heraclean template, especially in Prodicus’ parable transmitted by Xenophon (\emph{Mem}.~2.1), where the royal hero must choose the hard path of virtue over the easy path of vice.
The forms differ, but the threshold logic is constant: pre-eminence requires a decisive rejection of appetite, display, and domination.

\paragraph*{Hellenistic and Persian accession practice.}
Historical kings also meet ordeals just \emph{before} enthronement.
Cyrus is recognized through proved nobility after humble upbringing (Herodotus~1.114--122; Xenophon, \emph{Cyropaedia}).
Darius~I casts his first year in the Behistun inscription as defeating rival claimants to secure the crown.
Alexander seeks divine sanction at Delphi and Siwa before world conquest.
These are military-political and oracular trials rather than temptations, but the timing is the same: a testing immediately prior to rule.

\paragraph*{What is unique in the Gospel triad.}
(1) \textbf{Timing}: it stands \emph{immediately} after anointing and just before public kingship claims.
(2) \textbf{Setting}: the \emph{wilderness}, the classic chaos-zone of Egypt and Israel alike.
(3) \textbf{Form}: a \emph{threefold} offer that maps cleanly onto the ancient royal temptations catalog (\emph{bread/wealth}, \emph{spectacle/proof or pleasure}, \emph{kingdom/power}).
(4) \textbf{Resolution}: unlike Buddha (renounces kingship for enlightenment) or empire-builders (win by force), Jesus rejects all shortcuts and takes the path of sacrifice as the road to kingship.

\paragraph*{Conclusion.}
This scene is not a Mosaic pastiche and it is not a Buddhist import.
If the intent were to copy Moses, the setting would be Sinai with forty days before God, not a Satanic confrontation in the desert with a triadic escalation.
If the intent were to copy the Buddha, Māra’s daughters, demon armies, and enlightenment would appear, but they do not, and Buddhism is otherwise irrelevant to the Gospel’s narrative fabric.
What the text actually does is deploy the royal-ordeal topos that saturates the ancient world.

Diagnostic facts:
(1) \textbf{Timing:} it occurs immediately after anointing and immediately before public rule, i.e., at the accession threshold.
(2) \textbf{Setting:} the wilderness is the chaos-zone (Egypt’s \emph{isfet}, Israel’s testing ground), signaled by “with the wild beasts” (Mk~1:13).
(3) \textbf{Form:} the ordeal is a tight triad---bread (material provision), spectacle (public proof), kingdoms (sovereignty)---not generic suffering.
(4) \textbf{Scriptural grammar:} every answer quotes Deuteronomy’s wilderness-testing texts (Deut~8:3; 6:13; 6:16), tying Jesus’ trial to Israel’s forty-year pattern without reproducing Sinai’s scene.
(5) \textbf{Egyptian coronation logic:} pharaohs purify by abstinence, confront desert-chaos (\emph{isfet}) in ritual “smiting,” and receive regalia from the gods; the king proves himself in the chaos-border before enthronement.
(6) \textbf{Iranian/Indic triads:} Zarathustra refuses wealth, lust, power before inaugurating his mission; Siddhartha refuses desire, fearful spectacle, sovereignty before awakening; the categories match the Gospel’s triad even when the resolutions differ.
(7) \textbf{Hebrew royal ethics:} David rejects the shortcut to Saul’s throne; kingship comes by fidelity and ordeal, not grasping.
(8) \textbf{Hellenistic accession habit:} Cyrus, Darius, and Alexander face crises or oracular tests at the threshold of rule; the ordeal sits right before the crown.

Inference:
A Mosaic copy would preserve Sinai’s mountain theophany and omit Satan’s staged offers.
A direct Buddhist copy is unlikely as, although the philosophy of Jesus and Buddha share some affinities, we would have expected more Buddhist motifs at least somewhere in early Christian literature if there were direct influence.
The observed package---post-anointing \emph{wilderness} setting, Deuteronomic replies, \emph{triadic} royal offers, refusal of shortcuts, and launch of public kingship---fits the ancient \emph{royal trial} template exactly.

Therefore:
The evangelists are writing like royal historians, not like compilers of proof-texts or borrowers of exotica.
Egypt supplies the accession logic (chaos subdued before crowning).
Israel supplies the forty-day/forty-year testing and the words of reply.
Iran and India supply the triadic grammar of appetite, spectacle, and sovereignty.
Greece supplies the threshold habit of ordeal before rule.
The Gospels crystallize these strands into one accession scene: the anointed king enters the chaos-desert, refuses bread, spectacle, and kingdoms, and chooses the cross as the only legitimate path to rule.

\subsection{Possibly the most prominent element of the gospel of Matthew is the reference to the Jewish scriptures and the fulfillment of the prophecies at seemingly every turn.}\label{subsec:possibly-the-most-prominent-element-of-the-gospel-of-matthew-is-the-reference-to-the-jewish-scriptures-and-the-fulfillment-of-the-prophecies-at-seemingly-every-turn.}

As such the gospel seems to be with a very high probability written by a prominent Jewish priest, who was also a greek empire loyalist.
A speculative reconstruction, based on internal naming conventions and the Gospel's priestly concerns, has proposed identifying Matthew with a figure such as Mattathias ben Theophilus; this remains conjectural.
The proposal is not generally accepted because it requires a pre-70AD dating, but within the framework discussed in this book, that objection loses force, and the identification becomes at least worth considering alongside the traditional attribution.

\subsection{The gospel of Luke also includes a number of edits, most notably the birth narrative of Jesus.}\label{subsec:the-gospel-of-luke-also-includes-a-number-of-edits-most-notably-the-birth-narrative-of-jesus.}

The birth narrative seems to be full of internal contradictions and is at odds with the gospel of Matthew the events of which have some of the extremely remarkable facts corroborating to the rest of this theory.
The genealogy of Jesus in the gospel of Luke, notably has even more genuine figures from the Hasmonean dynasty down to the father of Mary, Heli, which is a short form of Eliakim, which is the same name as Joachim, which is the father of Mary in the protoevangelium of James.
If we accept that the gospel of Luke was written as a correspondence with the author of the gospel of Matthew, the dating of it would also not be far removed.

\subsection{The Gospel of Marcion deserves another look in this context.}\label{subsec:the-gospel-of-marcion-deserves-another-look-in-this-context.}

The gospel associated with Marcion (c. 140AD) survives only through hostile witnesses such as Tertullian (*Adversus Marcionem*, c. 207AD), Epiphanius (*Panarion*, c. 374AD), and the *Dialogue of Adamantius* (late 3rd/early 4th century).
Despite polemical framing, these sources preserve enough data to reconstruct key features of Marcion’s gospel.
It opened at Luke 3:1 (“In the fifteenth year of Tiberius”), contained no infancy narratives, omitted any clear prophecy of the 70AD temple destruction, and consistently presented simpler readings than the canonical Luke.

These features have long provoked debate.
Traditional church fathers insisted Marcion mutilated Luke, cutting whatever did not fit his theology.
But many modern textual critics --- from Harnack to more recent scholarship --- have put forward the alternative that Marcion’s gospel reflects an earlier edition of Luke.
The absence of infancy stories and later expansions is more naturally explained by an earlier recension than by wholesale excision.
The simpler language and alignment with “Western” text-type variants reinforce the sense of primitivity.

In terms of dating, this is significant.
If Luke was first written around 80--90AD, then Marcion’s text would require large-scale mutilation in 140AD.
But if Luke existed in multiple forms already, Marcion’s text is simply the form circulating in his communities of Asia Minor, while the longer canonical Luke represents a later expanded edition.

The key point is that the theory of Marcionite priority is not marginal speculation but a view seriously considered by major textual critics.
Rather than seeing Marcion as an innovator who created a gospel, it is far more economical to see him transmitting an earlier form of Luke.
The common misconception is that “Marcion wrote a gospel in 140AD,” when in reality he used a version already in circulation.
For the purposes of early dating, the critical lesson is that multiple Lukan recensions were alive in the first and early second centuries, and Marcion’s witness stands as one of the clearest windows into that diversity.
In the framework of an early Luke, this absence of a 70AD destruction motif makes perfect sense: the temple was still standing, and the author wrote before its fall, not after.
The same perspective explains why Marcion’s Pauline collection lacked the Pastoral Epistles.
These letters --- 1 and 2 Timothy and Titus --- all begin with explicit claims to Paul’s authority, yet modern scholarship is highly confident they are pseudonymous compositions from the early second century.
Their late addition served to domesticate Paul into a guardian of institutional order, while Marcion’s corpus preserved the authentic, radical Paul.
This reinforces the conclusion that Marcion’s library reflects an earlier, more original stage of both Luke and Paul, before later expansions and interpolations reshaped the tradition.
Luke’s own opening line --- “Many have undertaken to draw up an account before me” --- points directly to this situation.
Marcion’s gospel could represent precisely such an earlier account, with our canonical Luke being the edited and expanded form.
In that sense, the figure we call Luke may not be responsible for the longer recension preserved in the canon, but may instead have authored the shorter original that Marcion transmitted.
It is also entirely possible that Luke himself composed both a first version before 70AD and a second expanded recension after the fall of Jerusalem.
On this reading, Marcion simply had access to the earlier draft, while the canonical text reflects Luke’s later revision shaped by the traumatic context of the Temple’s destruction.
We also know that the author of Luke clearly wrote a sequel, the Acts of the Apostles, so it is very reasonable to suppose he could have composed both a first and second version of his gospel.
In this scenario his original version may have lacked a birth narrative simply because he had no source for it at the time, with such material only being added later in the expanded recension.

\subsection{The gospel of Luke describes how Jesus engaged with temple leaders and the Jewish elite when he was a child.}\label{subsec:the-gospel-of-luke-describes-how-jesus-engaged-with-temple-leaders-and-the-jewish-elite-when-he-was-a-child.}

We are told Joseph and Mary settled in Galilee, but they traveled to Jerusalem for the Passover festival every year.
This is consistent with Mary being a member of the Hasmonean dynasty, as the Hasmoneans were known to have a strong connection to the temple in Jerusalem.
This theory fully explains how Jesus would realistically be able to engage with the temple leaders and the Jewish elite at such a young age.
As per his wise man reputation, it is very likely that Jesus was a child prodigy, and would have been eager and able to engage with the temple leaders about scriptures and the law at a very young age.
Likely if Jesus was a child prodigy, he would have been studying Jewish and Greek scriptures and philosophy in Egypt and in Galilee.
Galilee libraries in Sepphoris would have been precisely well stocked with the works of the Greek philosophers and the Jewish scriptures fully accounting for Jesus's teachings.

\subsection{Luke is using greek historical writing style, and the gospel is written, not a religious text.}\label{subsec:luke-is-using-greek-historical-writing-style-and-the-gospel-is-written-not-a-religious-text.}

Ancient Greek Historians' Opening Formula: Luke's opening in Luke 1:1-4 mirrors a well-established convention in Greek historical writing.
The common formula in works by historians like Herodotus and Thucydides begins with a clear intention to provide a factual account and a claim of reliable sources.
The historian often says that he will provide a narrative based on careful research and interviews with eyewitnesses.
Similarly, Luke introduces his gospel by stating that many have already written accounts, and he aims to write ``an orderly account'' based on what he has investigated carefully (Luke 1:3).
This intro not only appeals to the authority of eyewitness testimony but establishes the writer's credibility---a trait highly valued in Greek historiography.

By establishing this historian's formula, Luke is distancing his account from mythical or legendary stories.
This is similar to the way Herodotus sets himself apart from storytellers or poets, aiming for historical accuracy.
His explicit claim of factual investigation is Greek historiographical technique.

He says many have undertaken to draw up an account of the things that have been fulfilled among us, just as they were handed down to us by those who from the first were eyewitnesses and servants of the word.
(note he was a servant of the word)

Emphasis on Historical Accuracy: Luke consistently anchors his narrative with real, known political figures and chronological details.
For example, he refers to the reign of Caesar Augustus (Luke 2:1), the governorship of Quirinius (Luke 2:2), and the rule of Pontius Pilate (Luke 23:1-25).
This is highly reminiscent of Greek historians, especially Thucydides, who grounds his narrative in precise dates and political actors (e.g., the years of the Peloponnesian War).
Public Acts and Decrees: The Roman census (Luke 2:1-3) is a key event that ties the story into the broader imperial political world.
This is a hallmark of Greek historical writing, where historians would use official decrees or military campaigns as turning points in the narrative.
Similarly, Acts uses references to imperial trials and Roman legal processes (e.g., Paul's trial before Gallio, Acts 18:12-17) as a means to legitimize Christianity and explain its spread through the empire.
Roman Emperors and Local Leaders: The mention of Herod the Great in the narrative helps place the events within a concrete political framework.
Like Polybius, who uses the actions of local leaders as windows into the larger workings of an empire, Luke does the same.
He mentions Herod as a regional figure whose reign is significant not just in its own right but as a piece of the greater political puzzle.

Luke's precise use of geographical locations, like Judea, Galilee, and Capernaum, adds to this historical framework, providing context for the actions of Jesus and the early church in a way that closely resembles the method of ancient historians who anchored their stories in known historical settings.

Unified Narrative: Luke and Acts together form a single, continuous narrative.
The structure itself is highly methodical, resembling the way historians like Polybius wrote histories that were divided into books or volumes that followed the chronological development of key events.
Luke's work is divided between the story of Jesus' life and ministry (Luke) and the early history of the Christian church (Acts).
The transition from the life of Jesus to the apostles' missions parallels how ancient historians would follow a ruler's life and then shift to his empire's activities.

Political and Religious Movements: Just as Polybius traces the rise and fall of Rome, or Suetonius traces the lives of emperors, Luke similarly focuses on political and religious figures and their movements through the empire.
His narrative of Paul's missionary journeys (in Acts) mirrors the structure of royal chronicles, detailing the challenges and legal trials faced by a prominent political figure (Paul) on his travels through the empire.

Speech as Political Tool: Luke often uses speeches by Jesus and Paul to express core theological and political ideas.
These speeches resemble the formal rhetoric found in Greek histories, where kings or political figures deliver speeches to clarify their goals and assert their legitimacy.
In Acts, for instance, Paul's speeches before kings (Agrippa, Acts 26) or the Sanhedrin (Acts 23) are presented in a historical context where the figure is asserting his political and religious identity.
Strategic Use of Legal Trials and Defenses: Much like Plutarch or Suetonius, who present key speeches in the trials of emperors or important political figures, Luke uses trials and defenses (such as Paul's before Festus and Agrippa) to emphasize the legitimacy of the Christian message.
These trials are strategically placed in the narrative to reflect not only political struggles but also theological defense---showing that the new movement aligns with or challenges imperial policy in ways that are deeply significant.

The Divine Favor of Jesus and Paul: The portrayal of Jesus' divine mission and Paul's apostolic authority mirrors the way Greek historians often depicted kings or emperors as possessing divine favor.
Luke's consistent depiction of Jesus' miracles and Paul's visions and divine encounters frames the early Christian movement as divinely ordained, much like the way kings and rulers were seen as divinely chosen or supported in Greek histories.

\subsection{Luke places the lineage of Jesus at the baptism, not at the birth.}\label{subsec:luke-places-the-lineage-of-jesus-at-the-baptism-not-at-the-birth.}

Scholars argue that Jesus lineage placed after birth narrative is an evidence that the birth narrative was added as a scribal edit.
However, this is not a valid argument, as if the scribe went to the effort of adding major piece of text and blend it well into the narrative and the core idea of the author, how could he have just omitted moving the lineage to the beginning of the text?

The obvious answer is that the lineage was not intended to be at the beginning of the text, and was given at the baptism of Jesus when proclaiming him as the rightful king.

\subsection{Later gospels and apocryphal texts}\label{subsec:later-gospels-and-apocryphal-texts}

It is important to note that to judge the significance of the dating of the texts, or really any claims in any historical work, it is important to address all evidence available, not just the evidence supporting the claim.
The emphasis is that while some texts like the gospel of Hebrews, the gospel of Marcion, the protoevangelium of James, gospel of Thomas, are indeed very likely very tied so independent source testimonies, there are a large number of writings, although seemingly expanded narrative, seem to be fully derived from other texts and do not bring any new source.
One example is the very common gospel of Nicodemus or the Acts of Pilate, which, although it adds a lot of human interactions, does not seem to bring up any fresh or contradictory information at all that would not be a natural embellishment or interpolation of the original texts.
Thus, it is very likely only composed based on other known texts and not on an unknown source or a direct testimony of first or second hand witness.

\subsection{Synoptic evidence for early dating.}\label{subsec:synoptic-evidence-for-early-dating.}

Mark's Greek is rugged and unpolished, which would be unusual if it had been written late by an established church.
Matthew shows knowledge of the Temple tax and practices, best explained if the Temple was still standing.
Luke uses refined Greek historiographical conventions, similar to Herodotus or Polybius, showing the training of a contemporary historian rather than a later theologian.
These features all point to first-generation dating.
If the gospels are first-generation documents anchored in the living memories of the participants and framed by Greek literary conventions, then the movement they describe is not a later theological construction but an organized political reality already active during Jesus' lifetime.
Recognizing this shifts the question from whether the early Christian movement arose gradually to how such a structure already existed in the very cities and networks where the gospels themselves emerged.
