This is the great mystery of faith:
Christ has died, Christ has risen, Christ will come again.

This is the core of Christian belief.
Christians taught love and forgiveness, and promised eternal life to all who believed in Christ.
That attracted many converts.
Then as the converts grew in number, the mystery was revealed.
Christ will come again.
Not just as a spiritual savior, but as the new emperor of the Greek world.
And so Greeks slowly grew the Christian community penetrating the Roman Empire, lower classes, urban centers, and the ruling elites.
Until such time that the Roman Empire itself was Christianized and the former Greek empire of Alexander the Great was restored as the kingdom of God on Earth.

A common unassailable but certainly wrong assumption of modern scholarship is that all early Christian literature was written in Greek because Greek was the lingua franca of the Roman empire.
According to this claim we would expect Julius Caesar, the People and the Senate of Rome, Virgil, Seneca to all be fluent and write and speak extensively in Greek.
In fact, it is correct Greek was the administrative language of nearly all of the former Greek empire which later became known as the Eastern Roman Empire.
Notably nearly all Christian writers including Clement of Rome, Ignatius of Antioch, Polycarp, Justin Martyr, Irenaeus, Origen wrote in Greek.
The first non-Greek church father is Tertullian, close to the end of the second century.
Christianity was a Greek-only religion in its literary and intellectual foundation: Jesus is mentioned in non-Greek sources, yet religious texts written in Coptic were not considered the same religion as Christianity.
The writings of the church fathers, read through the prism of the Byzantine resurgence, carry that expectation of restoration from the first generation to Constantine.

\section{The Phoenix: Imperial Symbolism in Early Christian Literature}\label{sec:phoenix-symbolism}

One of the earliest Christian writings outside the New Testament canon, \emph{1 Clement} (c.~96 AD), addressed from Rome to Corinth, contains an extended passage on the phoenix:

\begin{quote}
There is a bird which is called the Phoenix.
This is the only one of its kind and lives five hundred years.
When the time of its dissolution draws near, it makes for itself a coffin of frankincense and myrrh and other spices, and when the time is fulfilled it enters it and dies.
But as its flesh decays, a worm is produced, which is nourished by the moisture of the dead creature and puts forth wings.
Then, when it has grown strong, it takes up that coffin and flies from the land of Arabia to Egypt, to the city of Heliopolis, and, in the daytime, in the sight of all, it places itself on the altar of the sun.
\cite[25:1--5]{clement:firstclement}
\end{quote}

Modern scholarship almost invariably interprets this passage as an allegory of Christ's resurrection and of the general resurrection of the dead.
But this reading isolates the passage from the political context in which the phoenix already functioned as imperial iconography.
The phoenix was not a Christian invention; by the first century it was an established symbol of \emph{imperium renovatum}---``empire renewed.''

Roman imperial coinage from the Julio-Claudian period onward regularly paired the phoenix with legends like AETERNITAS AVG, SAECULUM NOVUM, or FEL TEMP REPARATIO.
Under Hadrian the phoenix appears radiate, standing on a laurel branch, as the reverse type of gold and silver issues advertising the ``eternity of the emperor'' (RIC II Hadrian 246--247).
Later, fourth-century coins of Constans and other Constantinian rulers show a phoenix on a rocky mound with the legend FEL TEMP REPARATIO, ``the happy restoration of the times,'' explicitly marking the renewal of imperial fortune after crisis.
In this Roman usage, the phoenix belongs to the emperor: it perches on his globe, stands beside his personified Aeternitas, or crowns his restoration of the age.

\emph{1 Clement}'s version of the story looks similar on the surface, but its geography is radically different.
The phoenix does not fly to Rome, or to the Capitoline, or to any imperial altar in Italy.
It carries the coffin with the bones of its ancestor from ``the region of Arabia into Egypt, to the city called Heliopolis,'' and there in broad daylight lays them on ``the altar of the sun'' before returning to its former dwelling.
That itinerary---Arabia $\rightarrow$ Heliopolis $\rightarrow$ solar altar---is not random mythic scenery.
It maps exactly onto the older Egyptian and Hellenistic phoenix traditions, which locate the bird's cult in Heliopolis and link it to the Bennu, the solar bird of Ra, ``lord of jubilees,'' venerated at the great sun temple of On \cite[2.73]{herodotus:histories}.
The phoenix's journey back to Heliopolis is a journey back to the cult-center of the Greek--Egyptian East, the former Ptolemaic heartland.

The contrast with imperial coinage is sharp.
On Roman money the phoenix is attached to the person of the Augustus and to Rome's own ``age renewed.''
In \emph{1 Clement} the same bird enacts a restoration that bypasses Rome entirely.
Its sacrifice takes place on a solar altar in Egypt, the symbolic center of the Greek East, not in the city of the Caesars.
Seen against this background, the passage is at minimum an appropriation, and arguably a quiet inversion, of Roman imperial phoenix ideology.
The sign of imperial eternity is re-told with an Eastern itinerary and relocated sacrifice.

Classical writers already describe the phoenix in imperial colors.
Pliny \cite[10.2]{pliny:nh} records that the bird's body is purple except the tail, and the Jewish-Alexandrian playwright Ezekiel the Tragedian (second century BC) \cite{ezekieldramatist:exagoge} describes its breast as \emph{porphyrous} (πορφυροῦς), ``purple,'' the same adjective used for imperial cloaks and royal textiles.
In Greek usage both πορφυροῦς and related terms like φοῖνιξ mark the deep red-purple associated with high status and royal rank.
Christian martyr literature picks up exactly this language: the \emph{Martyrdom of Polycarp} 15.2 describes the fire around Polycarp as ``like bread baking, or like gold and silver refined in a furnace,'' and 16.1 records that a great quantity of blood flowed from his body.
The phoenix, purple with royal plumage, and the martyr, clothed in the purple of blood, are drawn into the same semantic field of suffering, sovereignty, and cosmic renewal.

By the early third century the phoenix has become a standard Christian emblem of resurrection in both literature and art, but the imperial undertones remain.
Tertullian, in \cite[13]{tertullian:resurrection}, takes the phoenix as a ``most complete and unassailable symbol of our hope'' and finds it in Scripture itself: ``The righteous shall flourish like the phœnix.''
The verse is Psalm 92:12 (LXX 91:13), where the Hebrew has the palm tree, \emph{tamar} (\texthebrew{תָּמָר}); the Septuagint rendered it φοῖνιξ, the word that names both the palm and the bird, and Tertullian reads the bird: ``must men die once for all, while birds in Arabia are sure of a resurrection?''
Around the same time or slightly later, the Latin poem \cite{lactantius:phoenix}, generally attributed to Lactantius, elaborates the phoenix's thousand-year life, fiery death, and rebirth in a paradisical Eastern garden; while not overtly Christian, it was quickly read as an allegory of Christ's resurrection and of the paradisal homeland of the redeemed.
Phoenix-type resurrection imagery---typically a bird perched on a palm branch---appears in third-century funerary art, including the Catacomb of Priscilla (Greek Chapel complex), as a visual shorthand for eternal life.
Medieval descriptions of the Old St Peter's apse mosaic (attested in the \emph{Liber Pontificalis} tradition \cite{liberpontificalis} and in pre-demolition sketches recorded before the Renaissance rebuilding) describe a bird above the cross and tree-of-life motif, widely interpreted by art historians as a phoenix, visually transferring imperial ``eternity'' language into a Christological key.

Late antique imperial ideology did not stand still while this was happening.
From the fourth century on, Christian emperors themselves minted phoenix coins with explicitly restorative legends such as FEL TEMP REPARATIO, advertising ``happy times restored,'' and thereby fusing older pagan eternity symbolism with a Christianized rhetoric of providential renewal.
In that sense, imperial and Christian phoenix discourses cross: emperors appropriate the bird into their own Christian self-presentation, while bishops and theologians increasingly fold it into narratives of resurrection, martyrdom, and the Church's future glory.
By the Byzantine period the phoenix survives more as a literary and liturgical image than as a formal state emblem, but the underlying pattern is the same: it signals a polity or community that has passed through death and now stands again.

Read in this wider context, the phoenix allegory of \emph{1 Clement} is not an isolated curiosity.
It is the first point on a long Christian trajectory in which the phoenix consistently symbolizes resurrection and eternal life---in sermons, in Latin poetry, in catacomb art, in apse mosaics.
Early Christian catacombs display anchors, fish, dolphins, and tridents, images identical to those struck on Flavian imperial coinage.
These are Mediterranean imperial symbols rather than Jewish cultic signs.
What is distinctive in \emph{1 Clement} is that the phoenix's path runs not to Rome but to Heliopolis, and that its offering is made on the altar of the sun in Egypt rather than on a Roman imperial altar.
That choice of geography and cult location is transparent to anyone who knew the Heliopolitan phoenix traditions.
It encodes, within a seemingly benign nature illustration, a story about cosmic and political restoration centered in the Greek East.
The same bird that guarantees ``the eternity of the emperor'' on Roman coins here guarantees the renewal of a different order, returning to its own Eastern sanctuary and altar rather than to the city of the Caesars.
This is not merely a story about individual resurrection; it is an imperial renewal myth retold from the perspective of a Christian community that expects the age to turn and the East to rise again.

\section{Alexandria and Egypt Under Roman Rule: Hardship and Persecution}\label{sec:alexandria-egypt-suffered-enormous-hardship-and-persecution-during-the-roman-empire.}

The 200-year struggle of Rome to conquer all the Greek world ended in enormous tax burden on the newly conquered territories.
Almost all tax revenue of the Roman Empire at the time of Jesus and shortly after was coming from the newly conquered territories of the Greeks, and the single largest share came from Egypt, which sent more tribute to Rome than any other province (Josephus, \emph{War} 2.383--386).
It is often misunderstood that Egypt was simply so wealthy and more developed than the rest of the world that it could account for such a disproportionate share of the tribute.
Egypt was indeed the richest, but by no means by such a large margin.
The tax was simply so high to transfer the wealth of the Greeks to Rome and destroy the heartland of the Greek world and their capability to resist in the future.
Given the hardships that the Greeks suffered, there should be no doubt they would seek to restore their kingdom of God and not be ruled by the beast of Rome.
Egypt was considered property of the emperor (Tacitus, \emph{Annals} 2.59; Cassius Dio 51.17) and not subject to the normal senatorial and imperial governance system.
The suppression of Greek autonomy by Roman taxation created not only resentment but also a theological-political hope: that the empire of the Greeks, now crushed under Roman tribute, would rise again as the true kingdom of God.
This is why Christianity was entirely born in Greek literary forms---every epistle, every gospel, every apologetic treatise---because it belonged to the intellectual and political heirs of Alexander, not the heirs of Moses.
The Roman destruction of Jerusalem (70 AD) left Judaism fragmented, but the Greek cities of the East---Alexandria, Antioch, Ephesus, Smyrna---became the centers of Christian theology.

\section{The Eastern Mediterranean Synthesis: 300 Years of Unified Civilization}\label{sec:eastern-mediterranean-synthesis}

Before Alexander, the Eastern Mediterranean was a mosaic of distinct peoples with distinct gods.
Egyptians served Osiris, Isis, and a dense temple pantheon rooted in pharaonic kingship.
Israelite and Judean traditions spoke of El, Elohim, Yahweh, Baal, and Asherah in overlapping strands of worship and polemic.
Greeks honored Zeus, Athena, Apollo, Dionysus, and the Olympian company in city-based cults.
Phoenician cities revered Melqart, Astarte, and local Baals along the Levantine coast.
Syrian and Mesopotamian centers blended Hadad, Atargatis, Ishtar, and regional storm gods.
Each group had its own language, its own priesthood, its own cult calendar, and its own political institutions.

Alexander's conquests did not simply impose ``Greekness'' on these populations.
They forced sustained cohabitation and inaugurated a deliberate policy of fusion.
The new Macedonian rulers promoted one shared language, Greek, as the medium of administration, philosophy, and commerce.
They founded mixed cities in which Greek-style councils and assemblies governed populations that remained ethnically and religiously diverse.
The Ptolemies in Egypt adopted pharaonic titulature and presented themselves as successors to the ancient Egyptian kings.
The Seleucids in Syria and Mesopotamia issued decrees that combined Greek legal forms with local cult privileges.
Hellenistic rule thus created a common civilizational shell rather than erasing local substance.

From the beginning, the exchange was bidirectional.
Greeks who settled in Egypt encountered an older priestly culture and read it as a source of wisdom, not barbarism.
They identified Osiris with Dionysus (Herodotus 2.42, 2.144; Plutarch, \emph{De Iside et Osiride} 356a--b) and interpreted Egyptian myths in terms of their own philosophical categories.
Egyptian doctrines of post-mortem judgment, resurrection, and the weighing of the heart entered Greek discourse as venerable images of moral order.
Greeks who encountered Jews in Alexandria and the diaspora met a scriptural tradition that claimed one Creator God over all nations.
They heard the stories of Noah, Adam and Eve, Abraham, Joseph, Moses, and David as a universal moral history, not just the folklore of one tribe.

The Greek translation of the Hebrew scriptures in the third and second centuries BC, traditionally attributed to seventy-two scholars in Alexandria (Letter of Aristeas 301--311), marks a turning point in this process.
The Septuagint was produced by Jews, in Greek, for a Greek-speaking environment in which Jews and non-Jews shared the same language.
Once the Torah and then prophets, psalms, and wisdom books existed in Koine, they could circulate wherever educated readers moved.
These texts could be copied in the same scriptoria that preserved Homer and Plato.
They could be cited in philosophical argument and praised for their antiquity and moral rigor.
At the same time, Jews in Alexandria and other poleis adopted Greek rhetoric, philosophical vocabulary, and civic structures.
Synagogues functioned with Greek-style officials and assemblies, even when they preserved distinct liturgy and law.
Philo of Alexandria stands at the end of this trajectory, presenting Moses as the archetypal philosopher in a thoroughly Greek conceptual grammar \cite{philo:vimoysis}.

Egypt's contribution to the emerging synthesis was not limited to ritual and myth.
It offered a fully articulated theology of kingship in which the ruler embodied a higher divine reality.
It offered a structured eschatology in which individual lives were weighed against a standard of justice after death.
Such images made it natural to think of history, law, and rulership under the gaze of a single moral order.

Phoenician cities contributed the infrastructure that tied this world together.
Their harbors and shipping routes had linked Egypt, Syria, Cyprus, and the Aegean long before Alexander.
Their earlier consonantal script, centuries before the Hellenistic period, had influenced the development of the Greek alphabet (Herodotus 5.58--61).
By the first century, the everyday written medium in these ports was Greek, but the maritime network carrying Greek texts still rested on Phoenician foundations.

By the late Hellenistic and early Roman periods, the Eastern Mediterranean shared the same basic intellectual furniture.
Greek had become the common language of education from Alexandria to Antioch, from Ephesus to Corinth.
Urban elites across this zone learned the same canon of poets, historians, and philosophers.
Philosophical schools increasingly spoke of a single highest principle or god behind the many cults.
The Stoic logos, the Platonic One, and the God of Israel could be heard as different accents of one claim about the source of the cosmos.
Even where the old city gods remained in ritual practice, inscriptions and hymns began to invoke a ``Most High God'' above them.
The Hebrew scriptures depict God in bodily form, walking in Eden, wrestling Jacob, and showing Moses his back.
By the time of John's Gospel, God is invisible and known only through the Logos.
This shift reflects Platonic metaphysics and Philo's theology rather than the anthropomorphic language of the Hebrew Bible.

Epigraphic evidence makes this convergence visible.
Across Asia Minor, the Levant, and Cyprus, dedications to Theos Hypsistos, ``the Most High God,'' appear from the Hellenistic into the Roman imperial period.
These inscriptions are written in Greek and often come from non-Jewish environments.
They are typically aniconic, using altars and formulae rather than full cult statues.
They describe a single supreme deity in language that can echo both philosophical monotheism and Jewish biblical phrases.
They blur the boundary between Zeus, the philosophical first principle, and the God of Israel without simply identifying them.

This shared civilizational grammar was more durable than any single dynasty or sanctuary.
Egypt had passed from native to Persian to Macedonian to Roman control, yet its priestly wisdom survived each regime.
Athens had been defeated by Sparta in 404 BC and subordinated to Macedon in 338 BC, yet Greek paideia soon educated the children of its conquerors.
Jerusalem had been destroyed by Babylon in 586 BC, yet in exile Jews consolidated scripture and learned to be a people without a native monarchy.
Corinth had been sacked by Rome in 146 BC, then refounded as a Roman colony that quickly reasserted itself as a Greek-speaking commercial hub.
Alexandria had lost its last Ptolemaic ruler in 30 BC, yet Alexandrian scholarship and theology soon shaped imperial elites across the Mediterranean.

By the time of Jesus, the Eastern Mediterranean was no longer a patchwork of isolated ethnic cults.
It was a single oikoumene in which cities shared a language, an educational canon, and overlapping theological assumptions.
Local gods and traditions persisted, but they were now interpreted within a common framework that assumed one highest God above all.
It is within this deliberately fused and already unified civilization that the earliest Christian ecclesiae appear and make their claims about the true ruler of the world.
Celsus reports that Christians claimed Jesus was one among many divine figures who visited earth, with others arriving ``in groups of sixty or seventy at a time,'' a charge preserved in Origen's Contra Celsum \cite{origen:contracels}.
This testimony places Jesus within a wider Hellenistic pattern of divine emissaries rather than as a solitary heavenly intruder.
The claim only coheres within a polytheistic Greco-Roman framework in which multiple gods and agents moved between heaven and earth.

\section{What Happened Around the Year 70 AD?}\label{sec:what-exactly-happen-near-year-70-ad}

The year 70 AD is very critical in the study of early Christianity.
It is the anchor year on which the gospels are dated and the point where Christianity is thought to have changed from a failed apocalyptic movement into a spiritual religion.
But we need to ask why exactly this year is considered so decisive.
It was the year when the Temple in Jerusalem was destroyed by the Romans.
Yet all Christian writers of the time, including Paul, lived far away from Jerusalem and the Temple and did not maintain any real attachment to its priestly religion.
Therefore, the common assumption that the destruction of the Temple forced an entire theological shift in Christianity makes little sense.
It is not remotely plausible that this single event, in a city where almost none of the early authors lived, would have caused the wholesale rewriting of all gospels, epistles, and traditions, or the supposed burning of their original texts.

What we need to realize is that the destruction of the Temple in Jerusalem was only one of many violent events that marked the same decade.
The late 60s and early 70s AD were a time of massive instability across the Greek East of the Roman Empire.
In Galilee, Josephus fought the Romans as commander of the northern army in the \textit{First Jewish--Roman War} (66--73 AD).
At the same time, a fierce \textbf{Greek--Jewish civil war} broke out in Alexandria in 66 AD, when the Greek and Jewish quarters of the city erupted into mutual slaughter; the legions sent by Vespasian to restore order eventually razed much of the city \cite[2.497--507]{josephus:war}.
Similar \textbf{inter-ethnic riots} followed in Antioch and Damascus, where Greek and Jewish citizens massacred each other in the thousands \cite[7.43--45]{josephus:war}.
Meanwhile, across the wider Greek East, false claimants to the imperial throne appeared and local unrest destabilized the provinces \cite[2.8--9]{tacitus:histories}.
In Cyrene, Sicarii fugitives who fled after the fall of Jerusalem stirred further revolt (Josephus, \emph{War} 7.437--450), and Egypt remained under martial control throughout the 70s; the same tensions erupted again a generation later in the \textbf{Kitos War} (115--117 AD under Trajan).
That is essentially every major Greek city in the Roman Empire, all in revolt against Roman rule, all at the same time.
Even in Rome itself, several senators and equestrians were accused of complicity with eastern rebels and executed for treason.
This was not an isolated Jewish war---it was a general Greek and oriental rebellion against Roman rule, in which local ethnic conflicts and anti-imperial revolts fused into one massive civil breakdown.

The Roman high command itself saw these revolts as one system.
A preserved Tacitean tradition in Sulpicius Severus describes Titus' war council debating whether to spare or destroy the Temple.
Titus ultimately orders its destruction ``so that the religion of the Jews and of the Christians might be more completely overthrown,'' for both movements ``sprang from the same source, and if the root were destroyed, the branch would perish'' \cite[2.30.6--7]{sulpicius:chronica}.
This is not theology.
This is a Roman war document identifying Jews and the people who later acquire the name ``Christians'' as one political-religious bloc whose power rested on the same Jerusalem center.

And it was precisely at this date when historians put the best estimates for the execution of Paul (c.~64--67 AD) and Peter (c.~64--67 AD; Eusebius, \emph{Historia Ecclesiastica} 2.25) in Rome, and of James (c.~62 AD; Josephus, \emph{Antiquities} 20.200) in Jerusalem.
Paul and Peter were killed in the heart of the empire; James was killed in its most volatile province.
All three executions were political: Paul and Peter died under Nero's purge, and James was executed by the high priest Ananus during a power vacuum between Roman governors---an act so blatantly political that moderate Jewish leaders petitioned Rome to remove Ananus from office (Josephus, \emph{Antiquities} 20.201--203).

And here emerges the critical point.
Josephus, Paul, Peter, and James were not opponents on a fragmented religious landscape.
They operated inside the same political world: an imperial-restoration movement that drew Greek, Jewish, and eastern elites into a single anti-Roman horizon.
They did not share the same rituals or schools, but they served the same cause in different registers.
Their different ``religions'' were branches of the same royal cause, adapted to different audiences.
Josephus, in his \textit{Testimonium Flavianum} \cite[18.63--64]{josephus:ant}, describes Jesus as a leader who attracted many followers.
In the royal reading this statement is not paradoxical at all.
Josephus remained faithful to the religion of Moses, but he also recognized Jesus as a \textit{Christos}---a royal leader of Galilee---whose movement, though defeated, formed part of the wider rebellion that shook the entire Greek world around the year 70 AD.

\subsection{The Geographic Correlation: Ekklesia Network as Political Infrastructure}

The connection between Chapter 5's geographic findings and the year 70 AD revolts is inescapable.
Chapter 5 demonstrated that early Christian \emph{ekklesiai} appeared in \emph{every} major Greek city---Alexandria, Antioch, Ephesus, Sardis, Smyrna, Corinth, Philippi, Thessalonica---and in \emph{zero} major Latin-speaking cities.
This chapter shows that in the late 60s and early 70s AD, \emph{precisely those same cities}---Alexandria, Antioch, Ephesus, Sardis, and throughout Asia Minor---erupted in coordinated anti-Roman revolts.

This cannot be coincidence.

The issue is not simple correlation.
The cities that erupted in the 60s---Alexandria, Antioch, Ephesus, Sardis---are the same cities that formed the Hellenistic monotheist urban network across the Greek East.
The same shared ideology appears in every node: one God, ancestral law, rejection of Caesar's cult, and the expectation of a divinely sanctioned ruler.
The same shared institutional form appears in every node: Greek civic assemblies, councils, and voluntary associations---\emph{ekklesiai}---not village synagogues or tribal militias.
The same shared personnel appear across the entire crisis: Greek-educated monotheists, priests, philosophic teachers, urban elites, not isolated Judean zealots.
And the same chronology binds them all together: Alexandria 38 (Philo, \emph{In Flaccum} 36--96; \emph{Legatio} 120--137), Antioch in the 40s, Asia Minor in the 50s, Judea 66--73.
These are fronts of the same political world, not isolated accidents.

The \emph{ekklesia} network was not a religious association that happened to align with political boundaries; it was part of the same structural infrastructure that produced the revolt.
When we map the cities mentioned in Acts and the Epistles onto the cities that revolted in 66--73 AD, the overlap is nearly perfect.
The assemblies that Paul addressed in Corinth, Ephesus, Philippi, and Thessalonica were not passive religious communities; they were nodes in a restoration network that, when activated, launched a coordinated rebellion across the Greek-speaking East.

The statistical improbability of this pattern cannot be overstated.
If early Christian communities were simply missionary churches spreading by word-of-mouth evangelism, we would expect random geographic distribution, gradual diffusion across linguistic and political boundaries, and no correlation with political events.
Instead, we observe: (1) exclusive concentration in former Greek empire territories, (2) simultaneous appearance in all major Greek cities by the 60s AD, (3) coordinated revolts in those exact cities in 66--73 AD, and (4) immediate collapse of the network after the revolts were crushed, followed by (5) the restoration of the network only when imperial patronage shifted toward the East under Constantine.

This is not the pattern of a religious movement.
It is the pattern of a political restoration program using civic assemblies (\emph{ekklesiai}) as the organizational mechanism.

The overlap between the revolt map and the ekklesia map is not accidental and not merely correlation.
It reflects a single Hellenistic-monotheist infrastructure that stretched from Alexandria to Antioch to Asia Minor and finally to Jerusalem.
The assemblies that Paul encountered were not religious gatherings floating above politics.
They were the urban, Greek civic institutions through which every serious movement in the East---philosophic, political, or insurgent---operated.
When the decade of fire arrived, those same cities were the ones Rome feared, the ones that erupted, and the ones Titus explicitly identified with the same ideological root he sought to cut.
The \emph{ekklesiai} were not spectators to the revolt; they were part of the same world that produced it.
When Rome destroyed the Temple ``so that the religion of both Jews and Christians might be overthrown,'' it was not distinguishing denominations; it was crushing the political architecture of the entire Greek-monotheist East.

The ultimate success came three centuries later when the Eastern Roman (Byzantine) Empire was established as a Christian empire.

\section{Martyrdom Literature: The Universal Literature of Occupied Peoples}\label{sec:martyrdom-literature-occupied-peoples}

Modern scholarship treats the massive abundance of martyrdom literature in early Christianity as a religious phenomenon---evidence of faith, piety, and willingness to die for doctrine.
But this interpretation ignores the most obvious comparative context: martyrdom literature is the standard literary production of \emph{all} occupied peoples resisting imperial domination.

Wherever we examine occupied countries with strong national identity---Poland under partition, Vietnam under French colonial rule, Algeria under French occupation, Ireland under English domination---we find the same four elements: abundant martyrdom literature, a program of national liberation through moral superiority over the oppressor, the veneration of spiritual leaders and national heroes from both recent and distant history, and intensified religious organization centered on community gatherings.

The martyrs are celebrated not for personal holiness but as symbols of collective resistance.
Their deaths prove the nation still lives and will be restored.
But martyrdom alone is insufficient: the occupied people must also \emph{become better} than their oppressors---more virtuous, more disciplined, more worthy of sovereignty.
They must look to past leaders---ancient kings, prophets, warriors, and sages---as proof of former glory and models for restoration.
And they must gather regularly in churches, temples, community centers, or clandestine assemblies to maintain collective identity when direct political organizing is suppressed.
This is the standard program of national liberation movements: moral and civic transformation, inspired by heroic ancestors, sustained through religious community gatherings, as preparation for political restoration.

Poland, partitioned and occupied from 1795 to 1918, provides the clearest parallel.
Polish Romantic literature---especially the works of Adam Mickiewicz, Juliusz Słowacki, and Zygmunt Krasiński---created a massive corpus of martyrdom narratives that became foundational to Polish national identity.
The concept of Poland as the ``Christ of Nations'' (\emph{Chrystus narodów})---a nation that suffers crucifixion under foreign rule but will rise again---pervaded Polish poetry, drama, and prose throughout the occupation period.
This was not religious mysticism; it was political literature encoded in Christian imagery.
Martyrs like Tadeusz Kościuszko, executed or exiled for leading resistance, became secular saints whose suffering proved Poland's continued existence as a nation.
After independence was restored in 1918, the genre's political function largely ceased.

Vietnamese and Algerian resistance literature similarly framed their struggles as moral purification preparing the way for independence.
The pattern is universal: occupied peoples produce martyrdom narratives and preach self-improvement not as ends in themselves but as the path to national restoration.

Early Christian martyrdom literature fits this pattern exactly.
The \emph{Acts of the Martyrs}, the \emph{Martyrdom of Polycarp}, the \emph{Passion of Perpetua and Felicity}, and countless other texts were not primarily theological documents; they were political literature produced by an occupied people to maintain hope in restoration.
The martyrs died not for abstract doctrine but for refusing to acknowledge Roman imperial authority as ultimate.
They were executed for \emph{maiestas}---treason against the emperor---the same charge under which Paul, Peter, and James were killed.

The critical detail is that Christian martyrdom literature was produced almost exclusively in the Greek-speaking East, the very territories of the former Greek empire documented in Chapter 5.
Latin martyrologies appear much later and are demonstrably derivative.
The original martyrdom corpus comes from Smyrna, Antioch, Alexandria, Asia Minor---the cities that revolted in 66--73 AD and would later form the core of the Byzantine Empire.

This is not coincidence.
Martyrdom literature flourishes where political restoration remains a living hope.
When the Eastern Roman Empire was successfully Christianized under Constantine, the genre shifted: martyrs were no longer rebels against Rome but defenders of orthodoxy against heresy.
The political function of martyrdom---to sustain hope of liberation under occupation---disappeared once liberation was achieved.

The sheer volume of Christian martyrdom literature in the first three centuries is therefore not evidence of unusual piety but of sustained political resistance.
It is the literature of a people who believed their kingdom would be restored, who saw every martyr as proof that God had not abandoned them, and who eventually succeeded in establishing the Christian Byzantine Empire.
The martyrs were not dying for heaven; they were dying for the restoration of the kingdom of God to the Greek-speaking world.

\section{The Chi-Rho Symbol and Constantine's Eastern Imperial Restoration}\label{sec:chi-rho-constantinan-restoration}

The creation of the Eastern Roman Empire as a kingdom of God ruled by the Christ by a Christian army of Christian emperor Constantine was a very important event in the history of the world.
At that point the symbol of the soldiers was not a cross.
The symbol of the soldiers was the Chi-Rho symbol, which was a symbol of the Christos.
Note the shields only signified the Christos on them and not Jesus directly, which may have been a designation that ``we are the soldiers of the Christ,'' rightful king of the kingdom of God, and the battles for the restoration of the kingdom of the Christos.
If the war was thought for religious reasons primarily, there would have likely been a lot more emphasis on the cross and mass conversions to Christianity, and we do not see that being strongly emphasized in the historical records of this particular war.
We do however see the large emphasis on political changes and restoring the rule in the East.
Constantine's use of Chi-Rho on coins, shields, and banners was not a personal devotion to Jesus of Nazareth, but the creation of a Christic imperial monogram---a sign that the empire itself now belonged to the Christos, the anointed ruler.
Eusebius \cite[Book 1]{eusebius:vita} insists that the empire was renewed under the Christ, making it clear that Christianity functioned as the legitimizing religion of a new imperial order.

The timing is critical: Rome's Christianization coincided precisely with the division of the empire into Western and Eastern halves.
The restoration of the Eastern Roman Empire was the explicit goal of this movement, and Eusebius's writings confirm it.
He calls Constantinople the ``New Rome'' aligned with God's will, presenting the division of the empire and the Christianization of the East as a single, unified event---the fulfillment of the restoration theology that had animated early Christian writings for three centuries.

\section{Revelation: The Beast of Rome and the Kingdom Restored}\label{sec:revelation-political-restoration}

\emph{Note: Section~\ref{subsec:revelation-as-imperial-restoration-by-revolt-from-rome} treats Revelation as an operational document---the seven cities as an inspection circuit, the seals and trumpets as a campaign plan, the mark as counter-state administration.
This section treats Revelation as restoration theology: how its vision of Babylon's fall and New Jerusalem's descent fits the broader arc of Greek imperial hope.}

The Book of Revelation (c.~95 AD) is conventionally read as spiritual allegory about persecution and heavenly vindication.
But the political content is explicit and undeniable.
Revelation identifies Rome as ``the great city that rules over the kings of the earth'' (Rev 17:18), called ``Babylon'' in coded language to avoid direct sedition charges.
The beast with seven heads represents ``seven kings'' sitting on ``seven mountains'' (Rev 17:9--10)---Rome's seven hills make the identification unmistakable.
The whore of Babylon sits on ``many waters,'' explained as ``peoples, multitudes, nations, and languages'' (Rev 17:15)---the imperial \emph{oikoumene}.

The climactic vision is not spiritual escapism but political restoration:

\begin{quote}
Then I saw a new heaven and a new earth, for the first heaven and the first earth had passed away, and the sea was no more.
And I saw the holy city, new Jerusalem, coming down out of heaven from God, prepared as a bride adorned for her husband.
And I heard a loud voice from the throne saying, ``Behold, the dwelling place of God is with man. He will dwell with them, and they will be his people, and God himself will be with them as their God.'' (Revelation 21:1-3)
\end{quote}

This is not metaphor; it is restoration theology.
The ``new Jerusalem'' descending from heaven is the reestablishment of the kingdom of God on earth, replacing the beastly empire of Rome.
The vision describes a \emph{politeia}---a political order where God rules through his anointed king (\emph{Christos}) over a restored kingdom.
Revelation 11:15 makes this explicit: ``The kingdom of the world has become the kingdom of our Lord and of his Christ, and he shall reign forever and ever.''

This is imperial language.
The promise is not escape to heaven but the replacement of Roman imperium with divine imperium---exactly what occurred when Constantine established the Christian Byzantine Empire three centuries later.

\section{Origen of Alexandria (c.~185--254 AD): Universal Kingdom Through the Logos}\label{sec:origen-universal-kingdom}

Origen, writing from Alexandria---the intellectual center of the Greek East---articulated a vision of cosmic restoration through the Logos that merges Greek philosophy with Christian eschatology.
In \cite[8.68--75]{origen:contracels}, he explicitly defends Christianity against accusations of sedition by arguing that Christians seek not the overthrow of Rome by force but its transformation through divine truth.
But transformation toward what end?

Origen writes in \cite[8.68]{origen:contracels}:

\begin{quote}
We do say that there will eventually be, when the Word has prevailed over the entire rational creation and has converted every soul to His perfection, a time when He shall deliver up the kingdom to God the Father, when all rational beings shall have been moulded by the Word of God into one perfection.
\end{quote}

This is not otherworldly escapism.
Origen envisions a time when ``the Word has prevailed over the entire rational creation''---the \emph{oikoumene}, the civilized world.
The ``kingdom'' that Christ will ``deliver up to God the Father'' is a political reality: the universal kingdom unified under divine rule.

Origen's concept of \emph{apokatastasis} (ἀποκατάστασις)---the restoration of all things---is often spiritualized by modern interpreters, but his language is explicitly political.
In \emph{De Principiis}, he describes all things being ``re-established in a state of unity'' until ``God shall be all in all'' \cite[3.6.6]{origen:principiis}---building directly on Paul's vision of Christ subjecting all \emph{archai} (ἀρχαί, rulers) and \emph{exousiai} (ἐξουσίαι, authorities) to God (1~Cor~15:24), vocabulary drawn from imperial administration.
This is restoration theology: the reconstitution of a divinely ordered political cosmos, centered in the Greek-speaking East where Origen wrote.

\section{Irenaeus of Lyons (c.~130--202 AD): Providence and the Coming Kingdom}\label{sec:irenaeus-providence-kingdom}

Irenaeus, bishop of Lyons (a Greek-speaking city in Gaul), articulates a millennialist eschatology in \cite{irenaeus:advhaer} that envisions a literal earthly kingdom ruled by Christ.
Writing around 180 AD, Irenaeus interprets Daniel's vision of four kingdoms (Dan 2:31-45) as prophecy of the succession of empires---Babylonian, Median, Persian, and finally the Roman Empire---with the kingdom of God as the fifth and final kingdom that will crush and replace all earthly powers.

In \cite[5.26.1]{irenaeus:advhaer}, Irenaeus writes:

\begin{quote}
In the end, the stone cut without hands [from Daniel] shall smite the image upon the iron and clay feet and shall break them to pieces, and shall itself fill the whole earth.
This refers to the kingdom of our Lord...which shall break in pieces and bring to an end all kingdoms, and which shall itself endure forever.
\end{quote}

This is not spiritual allegory---it is explicit political prophecy.
The ``stone'' that smites the feet of iron and clay (Rome) is the kingdom of Christ, which will physically ``fill the whole earth'' and ``break in pieces...all kingdoms.''
Irenaeus interprets this as the establishment of a millennial kingdom on earth where the risen saints will reign with Christ \cite[5.32.1--5.35.2]{irenaeus:advhaer}, directly paralleling Revelation 20:4-6.

Crucially, Irenaeus's vision is geographically centered: the kingdom will be restored in Jerusalem and the surrounding lands, but Jerusalem here functions as the symbolic capital of the renewed oikoumene---the civilized world now under divine rule, not Roman imperium.

\section{Eusebius of Caesarea (c.~260--340 AD)}\label{sec:eusebius-of-caesarea-c.-260340-ad}

Eusebius tells the story of Christianity's triumph as restoration history.
He inscribes the tenth book of his church history as ``the perfect panegyric upon the restoration of the churches'' and titles its second chapter ``The Restoration of the Churches'' \cite[10.1--10.2]{eusebius:he}.
In his life of Constantine, the succession itself is a renewal: Constantine ``invested himself with his father's purple, and proceeded from his father's palace, presenting to all a renewal, as it were, in his own person, of his father's life and reign'' \cite[1.22]{eusebius:vita}.
The statue Constantine erected in Rome after his victory carries the same vocabulary; Eusebius records its inscription: ``I have preserved and liberated your city from the yoke of tyranny. I have also set at liberty the Roman senate and people, and restored them to their ancient distinction and splendor'' \cite[1.40]{eusebius:vita}.
Eusebius writes continuity into the past as well: in Historia Ecclesiastica 2.17 he identifies Philo's Therapeutae, described in De Vita Contemplativa, as early Christians and equates their communal ascetic life with Christian monasticism.
Philo uses the term \emph{monasterion} (μοναστήριον) for their individual dwellings, the direct linguistic source of the later Christian ``monastery.''
The institutional vocabulary of monasticism passes from a pre-Christian Hellenistic Jewish community into Christianity through this patristic identification.

\section{Victorinus of Pettau (c.~250--303 AD)}\label{sec:victorinus-of-pettau-c.-250303-ad}

Victorinus reads the Apocalypse as Roman political history.
Babylon he glosses directly: ``Babylon, that is, the Roman state'' \cite[Rev 7]{victorinus:apocalypse}.
The dragon's ``seven heads were the seven kings of the Romans'' \cite[Rev 12]{victorinus:apocalypse}.
The woman's seven hills he names outright: ``The seven heads are the seven hills, on which the woman sits. That is, the city of Rome'' \cite[Rev 17]{victorinus:apocalypse}.
On the seven kings of Revelation 17 he counts emperors: Galba, Otho, Vitellius, Vespasian, and Titus are the five who have fallen; ``One remains, under whom the Apocalypse was written---Domitian''; the one yet to come ``speaks of Nerva,'' who reigns briefly ``for he did not complete the period of two years''; and the beast that is of the seven is Nero, ``since before those kings Nero reigned'' \cite[Rev 17]{victorinus:apocalypse}.
And in this chronicle of emperors he finds a promise: the sealing angel ascending from the east comes ``for the restoration and establishment of the churches from the great and intolerable persecution'' \cite[Rev 7]{victorinus:apocalypse}.

The same commentary keeps returning to the imagery this chapter opened with---the sun, the purple, and the palm---six glosses in all.
Christ's face is like the sun ``on account of its rising and setting, and rising again, that He was born and suffered and rose again'' \cite[Rev 1]{victorinus:apocalypse}---the sun's setting and rising again as the figure of resurrection, the same logic that carries the phoenix of \emph{1 Clement} to the altar of the sun at Heliopolis.
The woman ``clothed with the sun'' likewise ``intimates the hope of resurrection and the glory of the promise'' \cite[Rev 12]{victorinus:apocalypse}.
The red dragon he recolors: it is ``of a red color---that is, of a purple color,'' a color his work gave him, ``for from the beginning\ldots he was a murderer'' \cite[Rev 12]{victorinus:apocalypse}.
On the sealed scroll he recalls Moses sprinkling the people with ``the wool purpled with the blood of the calf'' \cite[Rev 5]{victorinus:apocalypse}---the purple of blood again.
The elders' palms are victory tokens: ``the palm and the crown signify the same things, and these are not given save to the victor'' \cite[Rev 4]{victorinus:apocalypse}---and the palm, in Greek, is the φοῖνιξ.
The number of the beast he resolves into the sun's own name: the Greek letters τ, ε, ι, τ, α, ν---three hundred, five, ten, three hundred, one, and fifty---sum to six hundred and sixty-six and spell \emph{teitan} (τειταν), ``which the Gentiles call Sol and Phœbus'' \cite[Rev 13]{victorinus:apocalypse}; Irenaeus records the same reckoning and adds that ``even the sun is termed `Titan' by those who do now possess'' the rule \cite[5.30.3]{irenaeus:advhaer}.

\section{The Revolt Was Not Military: Rome Converted and the East Was Restored Peacefully}\label{sec:the-revolt-is-not-militaristic-rome-chose-to-spiritually-convert-to-christianity-and-the-eastern-roman-empire-was-restored-peacefully-while-the-christian-writers-started-to-praise-rome.}

Writing to the Roman magistrates in 197 AD, Tertullian names the charge against Christians: they refuse to sacrifice for the well-being of the emperor and so stand ``guilty of treason against a majesty more august'' \cite[28]{tertullian:apology}.
His defense is that Christians obtain what the sacrifice asks for, from the God who can grant it: ``We pray for life prolonged; for security to the empire; for protection to the imperial house; for brave armies, a faithful senate, a virtuous people, the world at rest'' \cite[30]{tertullian:apology}.
The empire's survival is the church's own concern, because the end of all things ``is only retarded by the continued existence of the Roman empire,'' so that in praying for its postponement Christians are ``lending our aid to Rome's duration'' \cite[32]{tertullian:apology}; the assembled congregation prays ``for the delay of the final consummation'' \cite[39]{tertullian:apology}.
He even claims the emperor himself: ``Caesar is more ours than yours, for our God has appointed him'' \cite[33]{tertullian:apology}.
Refusing the sacrifice while securing the emperor's safety by prayer is not rebellion; across the five passages quoted here---\emph{Apology} 28, 30, 32, 33, and 39---Tertullian makes one bid: to replace the imperial cult as the guardian of the empire.
Lactantius, writing shortly before Constantine's victory, declared in \cite[7.15]{lactantius:institutes} that Rome's destiny was to become Christian, and that the empire itself was God’s instrument to unify the world.
In \cite{augustine:civdei}, Augustine of Hippo (354--430 AD) offers a vision where the fall of the Roman Empire is viewed through a Christian lens.
He argues that the decline of the empire does not indicate the failure of divine providence.
While he focuses on the spiritual aspects of empire, he acknowledges the empire's role in preparing the world for the Christian kingdom, suggesting a ``restoration'' of the Roman Empire as a Christian entity in the future.

\section{The Shepherd of Hermas: The Most Popular Christian Text and the Nostalgia for Greek Glory}\label{sec:the-shepherd-of-hermas-c.-100-160-ad}

The \emph{Shepherd of Hermas} (c.~100--160 AD) was not merely one early Christian text among many---it was \emph{the most widely read Christian work} in the second and third centuries, more popular than most texts that later became canonical.
Manuscript evidence, patristic citations, and early Christian library catalogs demonstrate that \emph{Hermas} circulated more widely than the Gospels of Mark or John in many regions, was included in some early New Testament collections (Codex Sinaiticus), and was treated as Scripture by Irenaeus \cite[4.20.2]{irenaeus:advhaer}, Tertullian \cite{tertullian:marcionem}, Origen \cite[4.2.4]{origen:principiis}, and Clement of Alexandria \cite[1.29]{clement:stromata}.

This extraordinary popularity demands explanation.
Modern scholarship treats \emph{Hermas} as a bland moral allegory about repentance and church discipline, hardly the stuff of mass appeal.
But the political reading reveals why it resonated so powerfully with Greek-speaking audiences under Roman occupation: \emph{Hermas} is an extended allegory of imperial restoration, encoding Greek nostalgia for lost sovereignty in the imagery of an aging woman transformed back into youth, a ruined tower rebuilt to former glory, and a people purified and restored to their rightful place.

\subsection{The Vision of the Old Woman Transformed: Restoration of Greek Vigor}

The central vision of \emph{Hermas} \cite[Vision 1--4]{shepherd:hermas} features an old woman who progressively becomes younger.
In the first vision, she appears elderly, seated in a chair, holding a book.
In subsequent visions, she grows younger, her face more vigorous, until in the fourth vision she appears as a radiant bride.
Hermas asks the Shepherd who this woman is, and the answer given is: ``the Church.''

The text says the woman is ``the Church,'' but that is exactly how political allegory speaks in antiquity.
Plato hides civic blueprints under the language of the soul \cite{plato:republic}.
Aesop hides statecraft under animal fables with moral tags.
4 Ezra \cite{fourezdras} hides anti-Roman prophecy under visions explained as piety.
The Sibyllines \cite{sibyllineoracles} hide insurgent ideology under cosmic prediction.
Revelation hides imperial critique under the voice of Jesus.
The religious gloss is not the meaning; it is the camouflage that kept the meaning alive under Roman control.

But \emph{which} church, and \emph{what kind} of transformation?

Modern interpreters spiritualize this as the Church's moral renewal through repentance.
But the imagery is explicitly political: an old woman becoming young again is not moral improvement but \emph{restoration of youthful vigor}---precisely the language used in Greek and Roman political discourse for imperial renewal.
The Latin \emph{renovatio imperii} and Greek \emph{ananeōsis tēs arkhēs} (ἀνανέωσις τῆς ἀρχῆς) describe empire restored to its former strength after a period of decline.

The old woman represents the Greek-speaking world under Roman domination: aged, weakened, conquered.
Her transformation into a youthful bride represents the restoration of that world to its former glory---the re-establishment of Greek sovereignty, the renewal of Hellenistic imperial power.
This is not abstract ecclesiology; it is political allegory of the kind that circulated widely under occupation, where direct speech was dangerous but symbolic literature could encode resistance.

The woman's progressive rejuvenation mirrors the structure of resistance movements: first, the recognition of decline (the old woman, conquered and subjugated); then, the hope of restoration (her gradual youthing); finally, the vision of full restoration (the radiant bride).
This is the exact narrative arc of occupied peoples' literature: loss, resistance, and promised restoration.

\subsection{The Tower Vision: Rebuilding the Greek Oikoumene}

The most sustained allegory in \emph{Hermas} \cite[Vision 3, Similitude 9]{shepherd:hermas} is the vision of the tower being built.
The tower is constructed of stones, some of which are rejected, some of which require shaping and purification before they can be fitted into the structure.
The tower represents ``the Church,'' but the architectural imagery is unmistakably Greek: the tower (\emph{pyrgos}, πύργος) built of carefully fitted stones (\emph{lithoi}, λίθοι) echoes the monumental architecture of Hellenistic city-building.

Greek cities across the former Ptolemaic and Seleucid empires---Alexandria, Antioch, Ephesus, Pergamum---were distinguished by their monumental towers, gates, and colonnaded streets.
Under Roman rule, many of these structures fell into disrepair, their maintenance neglected as tribute flowed to Rome instead of local civic projects.
The vision of a tower being rebuilt, stone by stone, with careful selection and purification of materials, is a transparent allegory for the rebuilding of the Greek-speaking world: each stone a city or region, each act of purification a preparation for political restoration.

Critically, the stones that are rejected or discarded are those that cannot be fitted into the structure---those who refuse to participate in the restoration program.
This is not about individual salvation; it is about collective membership in a restored political body.
The language of purification and repentance (\emph{metanoia}) in \emph{Hermas} is civic, not merely moral: the call is to become worthy stones in the rebuilt tower, to prepare for the restoration of the kingdom.

\subsection{Why Was Hermas the Most Popular Christian Text?}

The extraordinary popularity of \emph{Hermas} in the Greek East becomes comprehensible once we read it as restoration allegory.
It was popular precisely because it articulated, in vivid and emotionally resonant imagery, the central political hope of occupied Greeks: that their world---aged and defeated---would be restored to youthful vigor, that their cities---ruined and neglected under Roman rule---would be rebuilt in glory, and that they themselves would be purified and made worthy of this restoration.

The text does not advocate immediate armed rebellion.
It preaches \emph{metanoia}---repentance, transformation, civic virtue---as the path to restoration.
This is the standard program of resistance movements under occupation: moral and civic renewal as preparation for political liberation.
The vision is gradualist, not revolutionary, but the end goal is explicit: the tower will be completed, the woman will be fully restored, and the faithful will reign in the renewed kingdom.

This is why \emph{Hermas} circulated more widely than most canonical texts in the second and third centuries.
It was not a theological treatise; it was the popular literature of Greek imperial restoration, written in allegory to evade Roman censorship but transparent to its intended audience.
The Greeks who copied, circulated, and venerated \emph{Hermas} understood exactly what it promised: the restoration of their kingdom, the rebuilding of their cities, and the vindication of their hope that God had not abandoned them to permanent subjugation.

When Constantine established the Christian Byzantine Empire in the fourth century, \emph{Hermas}' popularity began to wane.
The vision had been fulfilled: the tower was rebuilt as Constantinople, the old woman was restored as the youthful Christian empire, and the stones were fitted into place as the cities of the Greek East were reintegrated into a Christian imperial order.
The text had served its purpose, and its political function---to sustain hope under occupation---was no longer necessary.

\section{Clement of Alexandria's Exhortation to the Greeks (c.~190 AD)}\label{sec:clement-of-alexandrias-exhortation-to-the-greeks-c.-190-ad}

Clement opens the work with the legendary singers of Greek myth---Amphion who walled Thebes with music, Arion, and Orpheus who tamed the beasts---and sets the Logos against them as a ``new song'': ``Behold the might of the new song! It has made men out of stones, men out of beasts'' \cite[Chapter 1]{clement:exhortation}.
At the close of the work the Logos speaks in the first person: ``I desire to restore you according to the original model, that you may become also like Me'' \cite[Chapter 12]{clement:exhortation}.
The Greek frame was not only literary: Clement reserves his deeper teaching for the advanced, and says so in the open text: ``Some things I purposely omit, in the exercise of a wise selection, afraid to write what I guarded against speaking'' \cite[1.1]{clement:stromata}.
Eusebius preserves his account of where that teaching came from: after the resurrection the Lord imparted knowledge to James the Just, John, and Peter, ``and they imparted it to the rest of the apostles, and the rest of the apostles to the seventy'' \cite[2.1.4]{eusebius:he}.
That is not initiation into a secret; the chain has four named ranks---the Lord, the three, the apostles, the Seventy---and authority handed down through ranked office is the shape of a court, not a cult.
Origen, Basil of Caesarea, and Cyril of Jerusalem maintained the same hierarchical transmission of advanced doctrine.

\section{Cyprian of Carthage's The Lapsed (c.~250 AD)}\label{sec:cyprian-of-carthages-the-lapsed-c.-250-ad}

Cyprian, addressing a church just emerging from persecution, announces restoration in the past tense: ``Behold, beloved brethren, peace is restored to the Church; and although it lately seemed to incredulous people difficult, and to traitors impossible, our security is by divine aid and retribution re-established'' \cite[1]{cyprian:lapsed}.
The confessors return ``bearing the trophies from a prostrate enemy'' \cite[2]{cyprian:lapsed}.
Final judgment stays at the end of history: the merits of the martyrs will avail ``when, after the conclusion of this life and the world, His people shall stand before the tribunal of Christ'' \cite[17]{cyprian:lapsed}.
Restoration is not postponed to the end of the world; for Cyprian it has already happened inside history, while the tribunal waits.
